# Loops

## Loops in Pure JSON and YAML

Neither JSON or YAML support loops and therefore looping has to be built as a preprocessing layer in the engine.

## Looping Over Values

A simple looping strategy can be built into the engine as follows:

```json
{
    "loop": {
        "variable": "$l1$",
        "values": [1, 2, 3]
    },
    "content": {
        "val1": "$l1$",
        "val2": "Hello"
    }
}
```

and executed as follows:

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(loop=list(variable="$l1$", values=list(1,2,3)), content=list(val1="$l1$", val2="Hello")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

result <- list()
if (!is.null(inp$loop)) {
  var_name <- inp$loop$variable
  for (l in inp$loop$values) {
    var_value <- l
    loop_result <- purrr::modify_tree(inp$content, leaf = \(x) {
      if (is.list(x)) {
        x
      } else {
        gsub(var_name, var_value, x, fixed = TRUE)
      }
    })
    result <- append(result, list(loop_result))
  }
}
print(result)
```

## Nested Loops {#sec-loops-nested}

A more generic approach can be achieved by using `expand.grid()` to hold loop values.  An advantage of this approach is that nested loops can readily be constructed.  Take the example below with two loops:

```json
{
    "loop": [
        {
            "variable": "$loop1$", 
            "values": [1, 2, 3]
        },
        {
            "variable": "$loop2$", 
            "values": ["Alice", "Bob"]
        }
    ],        
    "content": {
        "val1": "$loop1$",
        "val2": "Hello $loop2$"
    }
}
```

The engine can be constructed using `expand.grid()` to build a data frame of all loop combinations and then running through them one by one is relatively simple:

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(loop=list(list(variable="$loop1$", values=list(1,2,3)), list(variable="$loop2$", values=list("Alice","Bob"))), content=list(val1="$loop1$", val2="Hello $loop2$")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

result <- list()

## list of all loop values and loop variable names
loops <- purrr::map(inp$loop, \(x) x$values) |>
    purrr::set_names(purrr::map_chr(inp$loop, \(x) x$variable))

## use expand.grid to build a data frame where each row contains a series of loop values
df_loops <- expand.grid(rev(loops), KEEP.OUT.ATTRS = FALSE)

## reconstruct a list with one entry for each series of loop values
l_loops <- split(df_loops, 1:nrow(df_loops))

for (loop_set in l_loops) {
  loop_updated <- inp$content
  for (loop_count in seq_along(loop_set)) {
    var_name <- names(loop_set)[loop_count]
    var_value <- loop_set[[loop_count]]

    loop_updated <- purrr::modify_tree(loop_updated, leaf = \(x) {
      if (is.list(x)) {
        x
      } else {
        gsub(var_name, var_value, x, fixed = TRUE)
      }
    })
  }
  result <- append(result, list(loop_updated))
}

print(result)
```

The data frame `df_loops` looks as follows:

```{r}
df_loops
```

## Building Loops from Variables

Using variables to build loops is a powerful construct since an entire output can be updated by simply changing the values in a variable or pointing to a different one.  We can achieve it by using two passes - one to identify and substitute variable values and one to build the loops.  The example below follows the same code as @sec-loops-nested but replaces *Alice* and *Bob* in `$loop2` values with a variables (`$people$`) that points to *Alice* and *Bob*.  Code explanation will follow the example.

```json
{
    "variables": {
        "people": ["Alice", "Bob"]
    },
    "loop": [
        {
            "variable": "$loop1$", 
            "values": [1, 2, 3]
        },
        {
            "variable": "$loop2$", 
            "values": "$people$"
        }
    ],        
    "content": {
        "val1": "$loop1$",
        "val2": "Hello $loop2$"
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(people=list("Alice","Bob")), loop=list(list(variable="$loop1$", values=list(1,2,3)), list(variable="$loop2$", values="$people$")), content=list(val1="$loop1$", val2="Hello $loop2$")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

## parse variables
inp_var_replace <- inp

for (v in names(inp$variables)) {
  var_name <- paste0("$", v, "$")
  var_value <- inp$variables[[v]]
  inp_var_replace <- purrr::modify_tree(
    inp_var_replace, 
    leaf = \(x) {
      if (is.list(x)) {
        x
      } else if (is.character(var_value)) {
        gsub(var_name, var_value, x, fixed = TRUE)
      } else if (identical(x, var_name)) {
        var_value
      } else {
        x
      }
    }
  )
}

## execute loops
result <- list()
loops <- purrr::map(inp_var_replace$loop, \(x) x$values) |>
    purrr::set_names(purrr::map_chr(inp_var_replace$loop, \(x) x$variable))
df_loops <- expand.grid(rev(loops), KEEP.OUT.ATTRS = FALSE)
l_loops <- split(df_loops, 1:nrow(df_loops))


for (loop_set in l_loops) {
  loop_updated <- inp_var_replace$content
  for (loop_count in seq_along(loop_set)) {
    var_name <- names(loop_set)[loop_count]
    var_value <- loop_set[[loop_count]]

    loop_updated <- purrr::modify_tree(loop_updated, leaf = \(x) {
      if (is.list(x)) {
        x
      } else {
        gsub(var_name, var_value, x, fixed = TRUE)
      }
    })
  }
  result <- append(result, list(loop_updated))
}

print(result)
```

We can split the code into two sections: replacing variables and running loops.  The initial json input looks like this:

```json
{
    "variables": {
        "people": ["Alice", "Bob"]
    },
    "loop": [
        {
            "variable": "$loop1$", 
            "values": [1, 2, 3]
        },
        {
            "variable": "$loop2$", 
            "values": "$people$"
        }
    ],        
    "content": {
        "val1": "$loop1$",
        "val2": "Hello $loop2$"
    }
}
```

The loop variable `$loop2$` points to a variable, `$people$` which holds the values *Alice* and *Bob*.  When we run the initial part of the code (*parse variables*) we generate a new version of *inp* called *inp_var_replace* in which any variable used in inp$variables is replaced with its value.  Since we have converted the JSON format to a list in R, we can readily replace a variable name with a list of values.  The code snippet is shown below, followed by the resulting *inp_var_replace*, shown in JSON format for comparison to the original.

```{r}
#| eval: false
## parse variables
inp_var_replace <- inp

for (v in names(inp$variables)) {
  var_name <- paste0("$", v, "$")
  var_value <- inp$variables[[v]]
  inp_var_replace <- purrr::modify_tree(
    inp_var_replace, 
    leaf = \(x) {
      if (is.list(x)) {
        x
      } else if (is.character(var_value)) {
        gsub(var_name, var_value, x, fixed = TRUE)
      } else if (identical(x, var_name)) {
        var_value
      } else {
        x
      }
    }
  )
}
```

```json
{
    "variables": {
        "people": ["Alice", "Bob"]
    },
    "loop": [
        {
            "variable": "$loop1$", 
            "values": [1, 2, 3]
        },
        {
            "variable": "$loop2$", 
            "values": ["Alice", "Bob"]
        }
    ],        
    "content": {
        "val1": "$loop1$",
        "val2": "Hello $loop2$"
    }
}
```

When the second part of the code (*execute loops*) is run, the loop values of *Alice* and *Bob* are used for loop variable `$loop2$` producing the desired output.  By simply changing the values in `$people$`, the output can be affected without having to change values in individual loops.  This can be especially effective when the variable contents are looped over multiple times.
