# Variables

## Variables in Pure JSON and YAML

JSON does not have a concept of variables or variable reuse.  YAML does include anchors (&) and aliases (*) which allow for reuse.  For example, in YAML the following is permitted:

```yaml
threshold: &thresh 30

filters:
  - col: mpg
    operator: ">"
    value: *thresh
```

`&thresh` defines an anchor and `*thresh` reuses that value.  There are limitations to how anchors and aliases are used.  They are pure substitution, not symbolic evaluation, meaning that the following is not valid:

```yaml
threshold: &thresh 30
max_threshold: *thresh + 10
```

## Building a Variable Parser

### Simple Case {#sec-variable-simple}

To encode variables in JSON format, consider the following:

```json
{
    "variables": {
        "myVar1": 1,
        "myVar2": "Hello"
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(myVar1=1, myVar2="Hello")), 'data/myfile.json', auto_unbox = TRUE)
```

When imported into R, this JSON file appears as:

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
inp
```

The variables listed under `variables` in the JSON file are available through `inp$variables$name` where `name` is the name of a variable.

Now all we need is a method to replace a variable name with the value of the variable.  This can be achieved by defining a syntax for variables.  For example, in the case below we define variable use as `$name$` where `name` is the name of a variable.  Here `subValue` is set to be equal to the variable `myVar2` by defining it as `$myVar2$`.

```json
{
    "variables": {
        "myVar1": "Hello",
        "myVar2": "World"
    },
    "subValue": "$myVar1$"
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(myVar1="Hello", myVar2="World"), subValue="$myVar1$"), 'data/myfile.json', auto_unbox = TRUE)
```

Finally, the engine requires a method to replace variables in the JSON prior to the JSON contents being interpreted and processed.

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
if (!is.null(inp$variables)) {
  for (v in names(inp$variables)) {
    var_name <- paste0("$", v, "$")
    var_value <- inp$variables[[v]]
    inp$subValue <- gsub(var_name, var_value, inp$subValue, fixed = TRUE)
  }
}
print(inp)
```

Since we are simply replacing text with values, variables can be placed as needed.  For example:


```json
{
    "variables": {
        "myVar1": "Hello",
        "myVar2": "World"
    },
    "subValue": "$myVar1$ there"
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(myVar1="Hello", myVar2="World"), subValue="$myVar1$ there"), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
if (!is.null(inp$variables)) {
  for (v in names(inp$variables)) {
    var_name <- paste0("$", v, "$")
    var_value <- inp$variables[[v]]
    inp$subValue <- gsub(var_name, var_value, inp$subValue, fixed = TRUE)
  }
}
print(inp)
```

### Extending to Multiple Replacements {#sec-variable-multiple}

The example in @sec-variable-simple demonstrated a simple case of variable parsing but how can this work in a more generalized approach when the JSON structure we wish to process is more complex.  One way to tackle this is to use `purrr::modify_tree()` which modifies the contents of a nested list.  An example if its use for variable parsing is shown below.

::: {.callout-tip}
`purrr::modify_tree()` recursively walks through a nested list and applies a function to either the *leaves* (terminal values), the *nodes* (lists that contain other elements), or both, returning a new list with the same structure.
:::

```json
{
    "variables": {
        "myVar1": "Hello",
        "myVar2": "World"
    },
    "contents": {
        "val1" : "$myVar1$ there",
        "val2" : "$myVar2$",
        "val3": {
            "val4": {
                "nested": "data",
                "nested_var": "$myVar1$ $myVar2$"
            },
            "val5": [
                "Regular text",
                "$myVar1$ $myVar2$"
            ]
        }
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(myVar1="Hello", myVar2="World"), contents=list(val1="$myVar1$ there", val2="$myVar2$", val3=list(val4=list(nested="data", nested_var="$myVar1$ $myVar2$"), val5=list("Regular text", "$myVar1$ $myVar2$")))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
contents <- inp$contents

for (v in names(inp$variables)) {
  var_name <- paste0("$", v, "$")
  var_value <- inp$variables[[v]]
  contents <- purrr::modify_tree(
    contents, 
    leaf = \(x) {
      if (is.list(x)) {
        x
      } else {
        gsub(var_name, var_value, x, fixed = TRUE)
      }
    }
  )
}

print(jsonlite::toJSON(contents, auto_unbox = TRUE, pretty = TRUE))
```

Let's unpack what's happening.  `contents` can be considered as a tree, containing both nodes (branches that contain other elements) and leaves (values that sit at the end of branches), such that:

```
contents                                    (node)
├─ val1 : "$myVar1$ there"                  (leaf)
├─ val2 : "$myVar2$"                        (leaf)
└─ val3                                     (node)
   ├─ val4                                  (node)
   │  ├─ nested      : "data"               (leaf)
   │  └─ nested_var  : "$myVar1$ $myVar2$"  (leaf)
   └─ val5                                  (node)
      ├─ [1] "Regular text"                 (leaf)
      └─ [2] "$myVar1$ $myVar2$"            (leaf)
```

We apply `purrr::modify_tree()` to `contents` in order to perform variable substitution within a deeply nested JSON structure.  The loop iterates over each entry in `inp$variables` and for each variable the following is performed:

1. the variable name is converted into a placeholder, eg `myVar1` becomes `$myVar1$`.
2. The corresponding replacement value is retrieved, eg `Hello`.
3. Within `modify_tree()`:
    - If the element is a list then it represents an internal node of the tree so simply return it unchanged.
    - If the element is not a list then it must be a leaf value, eg a string or numeric.  In this case `gsub` is used to replace any occurrence of the variable with its value.

This process is repeated for each variable, so that all placeholders, no matter how deeply nested, will be replaced with their corresponding values.  Once complete, `contents` holds the updated list with the original structure.

### Extending to Different Types of Data {#sec-variables-different}

The concept of variables can be extended to account for different types of data.  The approach in @sec-variable-multiple works well for strings but fails in the cases of numeric, boolean and vectors.  To extend for numerics and booleans we cannot use `gsub()` and have to check for equivalence as shown in the example that follows:

```json
{
    "variables": {
        "myVar1": true,
        "myVar2": 55
    },
    "contents": {
        "val1" : "$myVar1$",
        "val2" : "$myVar2$"
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(myVar1=TRUE, myVar2=55), contents=list(val1="$myVar1$", val2="$myVar2$")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
contents <- inp$contents

for (v in names(inp$variables)) {
  var_name <- paste0("$", v, "$")
  var_value <- inp$variables[[v]]
  contents <- purrr::modify_tree(
    contents, 
    leaf = \(x) {
      if (is.list(x)) {
        x
      } else if (is.character(var_value)) {
        gsub(var_name, var_value, x, fixed = TRUE)
      } else if (identical(x, var_name)) {
        var_value
      } else {
        x
      }
    }
  )
}

print(jsonlite::toJSON(contents, auto_unbox = TRUE, pretty = TRUE))
```

The `modify_tree()` function now runs through the following checks:

-  If `x` is a list then return `x` (`x` is a node, not a leaf).
-  If `x` is a character vector then attempt to substitute the variable.
-  If `x` is not a character vector but is identical to a variable name, replace it with the variable value.
-  Otherwise return `x`.

This approach works when direct replacement is all that is required (eg replacing `$myVar2` with `55`).  It will not work when a variable is used in an expression (eg `$myVar2$ + 10`).

## Expression Parsing

In order to work with expressions such as `$myVar2$ + 10` we need to build an expression parser.  When evaluating expressions there is a temptation to use `eval(parse())` directly.  This should be avoided as it could lead to arbitrary code execution.  A safe expression parser for data wrangling has already been discussed in @sec-expression-parser and it can modified for expressions containing variables.  Safe expression handlers parse an expression into an AST (abstract syntax tree) and validate it before evaluation.

Let's start with a JSON input:

```json
{
    "variables": {
        "myVar1": true,
        "myVar2": 55
    },
    "contents": {
        "val1": "$myVar1$",
        "val2": "$myVar2$ + 10",
        "val3": "($myVar2$ > 50) && $myVar1$"
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(myVar1=TRUE, myVar2=55), contents=list(val1="$myVar1$", val2="$myVar2$ + 10", val3="($myVar2$ > 50) && $myVar1$")), 'data/myfile.json', auto_unbox = TRUE)
```

The engine requires several components, which will be built as a group of functions.  The approach of variable substitution followed by expression evaluation can be summarized by the following pipeline:

```{mermaid}
flowchart TD
    A(Substitute Variables) --> B(Detect Expression)
    B --> C(Parse)
    C --> D(Validate)
    D --> E(Evaluate)
```

### Variable Substitution

First we substitute the variables using a slightly different engine to the one in @sec-variables-different.  The previous substitution approach is too simple for expression building.  The following situations need to be solved:

-  non-character variables were only replaced if they matched exactly (eg `$myVar2$` would replace in `$myVar2$` but not in `$myVar2$ + 1`).  An updated substitution code requires numeric and logical variables to be substituted inside strings.
-  change `as.character()` approach.  This works for simple value numeric and logical replacements but fails for vectors and ranges.  The base function `deparse()` is a better approach for expression building.

::: {.callout-note}
## Why `deparse()`?
`deparse()` converts values back into source code, which is what is needed for expression parsing.  The difference between `as.character()` and `deparse()` is shown below:

| value | `as.character()` | `deparse()` |
| --- | --- | --- |
| `2` | "2" | "2" |
| `TRUE` | "TRUE" | "TRUE" |
| `c(1, 2)` | "1" "2" | "c(1, 2)" |
| `1:3` | "1" "2" "3" | "1:3" |
:::

```{r}
substitute_vars <- function(x, vars) {
  if (!is.character(x)) return(x)

  for (v in names(vars)) {
    replacement <- if (is.logical(vars[[v]]) || is.numeric(vars[[v]])) {
      deparse(vars[[v]])
    } else {
      as.character(vars[[v]])
    }

    x <- gsub(paste0("$", v, "$"), replacement, x, fixed = TRUE)
  }

  x
}
```

```{r}
vars <- list(myVar1 = "text string", myVar2 = 22)
substitute_vars("this is a $myVar1$", vars)
substitute_vars("$myVar2$ + 10", vars)
```

### Parsing and Evaluating Expressions

After substituting variables, the next steps are:

1. **Parse** the expression into an Abstract Syntax Tree (AST) without executing.
2. **Validate** the AST to ensure only allowed operators are present.
3. **Evaluate** in a restricted environment with only necessary operators.

This approach has been discussed in detail in @sec-expression-handling.  The functions outlined in this chapter, along with the wrapper function for safe expression handling presented in @sec-validate-function can be used to evaluate variable expressions.

### A function to run it all

Now we have all the components, we just need a function to run variable substitution followed by validation and evaluation.  Here we extend the `substitute_vars()` function to include the additional steps in a simple helper function.

```{r}
substitute_and_eval <- function(x, vars) {
  if (!is.character(x)) return(x)

  # Substitute variables
  for (v in names(vars)) {
    replacement <- if (is.logical(vars[[v]]) || is.numeric(vars[[v]])) {
      deparse(vars[[v]])
    } else {
      as.character(vars[[v]])
    }

    x <- gsub(paste0("$", v, "$"), replacement, x, fixed = TRUE)
  }

  # Evaluate expressions
  tryCatch(
    {
      eval_safe_expr(x, mode = "evaluate", eval_context = "scalar")
    },
    error = function(e) {
      # Return untouched
      x
    }
  )

}
```

`substitute_and_eval()` is the function run within `purrr::modify_tree()` to substitute and evaluate for each leaf of our tree.  The function takes a named list of variables and a list or character string and performs the following steps:

-  variable substitution - substitute only known variable names
-  expression validation - check that only desired functions/operators are used in the expression
-  safe evaluation - evaluated in a safe environment
-  fallback - if not an expression, then return unchanged

### Putting it all Together

Our final, complete code is below.  It consists of:
-  a list of allowed expressions
-  the validator (a data-masked expression validator from @sec-expression-handling is also included for completeness)
-  a safe evaluation environment
-  wrapper function to parse, validate and evaluate
-  function to substitute variables and call wrapper function to parse, validate and evaluate resulting expression

```{r}
# List of permitted operators
# Any other operator or function will throw an error during validation
allowed_operators <- c("+", "-", "*", "/", "^", 
                       ">", ">=", "<", "<=", "==", "!=", 
                       "&&", "||", "!", "c", "mean", "sd",
                       "&", "|")

# Function to validate scalar expression                       
validate_var_expr <- function(expr) {
  
  # Atomic values (numbers, strings, logicals) are always safe
  if (rlang::is_atomic(expr)) {
    return(TRUE)
  }

  # Check function calls
  if (rlang::is_call(expr)) {
    fn <- rlang::call_name(expr)

    # Parentheses are grouping operators, not function calls
    if (fn == "(") {
      validate_var_expr(expr[[2]])
      return(TRUE)
    }

    # Reject disallowed operators
    if (!fn %in% allowed_operators) {
      stop("Disqualified operator: ", fn, call. = FALSE)
    }

    # Recursively validate arguments
    for (i in 2:length(expr)) {
      validate_var_expr(expr[[i]])
    }

    return(TRUE)
  }

  # Anything else is rejected
  stop("Unsupported expression type", call. = FALSE)
}

# Function to validate data-masked expression
# Included for completeness (not used when evaluating scalar expressions)
validate_expr <- function(expr) {
  
  # Atomic values (numbers, strings, logicals) are always safe
  if (rlang::is_atomic(expr)) {
    return(TRUE)
  }

  # Symbols (variables and column names) are safe
  if (rlang::is_symbol(expr)) {
    return(TRUE)
  }

  # Check function calls
  if (rlang::is_call(expr)) {
    fn <- rlang::call_name(expr)

    # Parentheses are grouping operators, not function calls
    if (fn == "(") {
      validate_expr(expr[[2]])
      return(TRUE)
    }

    # Reject disallowed operators
    if (!fn %in% allowed_operators) {
      stop("Disqualified operator: ", fn, call. = FALSE)
    }

    # Recursively validate arguments
    for (i in 2:length(expr)) {
      validate_expr(expr[[i]])
    }

    return(TRUE)
  }

  # Anything else is rejected
  stop("Unsupported expression type", call. = FALSE)
}

# Build a safe environment to evaluate scalar expressions
safe_env <- rlang::new_environment(
  data = list(
    `+`  = `+`,
    `-`  = `-`,
    `*`  = `*`,
    `/`  = `/`,
    `^`  = `^`,
    `>`  = `>`,
    `<`  = `<`,
    `>=` = `>=`,
    `<=` = `<=`,
    `==` = `==`,
    `!=` = `!=`,
    `&`  = `&`,
    `|`  = `|`,
    `&&` = `&&`,
    `||` = `||`,
    `!`  = `!`,
    `(` = `(`,
    c = c,
    mean = mean,
    sd = sd,
    `&` = `&`,
    `|` = `|`
  )
)

# Wrapper function for parsing and evaluation
eval_safe_expr <- function(
  expr_input,
  mode = c("inject", "evaluate"),
  eval_context = c("data_mask", "scalar"),
  data = NULL,
  validator = NULL
) {

  mode <- match.arg(mode)
  eval_context <- match.arg(eval_context)

  # Parse if a string is supplied
  expr <- if (is.character(expr_input) && length(expr_input) == 1) {
    rlang::parse_expr(expr_input)
  } else {
    expr_input
  }

  # If validator is not specified then choose it based on eval_context
  if (is.null(validator)) {
    validator <- if (eval_context == "data_mask") {
      validate_expr
    } else {
      validate_var_expr
    }
  }

  # Validate
  # Will stop on failure
  validator(expr)

  # mode: inject
  # Return validated expression
  if (mode == "inject") {
    return(expr)
  }

  # mode: evaluate
  # Evaluate the expression
  if (eval_context == "data_mask") {
    # Evaluate data mask using eval_tidy
    if (is.null(data)) stop("Data Required")
    return(rlang::eval_tidy(expr, data = data))
  } else {
    # Safely evaluate scalar expression
    eval(expr, envir = safe_env)
  }
}

# function to substitute variables and evaluate resulting expression
substitute_and_eval <- function(x, vars) {
  if (!is.character(x)) return(x)

  # Substitute variables
  for (v in names(vars)) {
    replacement <- if (is.logical(vars[[v]]) || is.numeric(vars[[v]])) {
      deparse(vars[[v]])
    } else {
      as.character(vars[[v]])
    }

    x <- gsub(paste0("$", v, "$"), replacement, x, fixed = TRUE)
  }

  # Evaluate expressions
  tryCatch(
    {
      eval_safe_expr(x, mode = "evaluate", eval_context = "scalar")
    },
    error = function(e) {
      # Return untouched
      x
    }
  )

}
```

and can be used as follows:

```json
{
    "variables": {
        "myVar1": true,
        "myVar2": 55
    },
    "contents": {
        "val1": "$myVar1$",
        "val2": "$myVar2$ + 10",
        "val3": "($myVar2$ > 50) && $myVar1$"
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(variables=list(myVar1=TRUE, myVar2=55), contents=list(val1="$myVar1$", val2="$myVar2$ + 10", val3="($myVar2$ > 50) && $myVar1$")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

result <- purrr::modify_tree(
  inp$contents, 
  leaf = \(x) substitute_and_eval(x, inp$variables)
)
print(result)
```
