# Selection and Ordering

## Selecting Columns

### `dplyr::select()`

`dplyr::select()` picks or drops columns based on their names or positions.  It returns a new data frame containing only the chosen columns.  The number of rows is not affected.

### Retaining Columns

`dplyr::select()` takes one or more expressions separated by commas to specify columns.  For example, below the columns *mpg*, *cyl* and *wt* are retained, whilst all other columns are dropped.

```{r}
dplyr::select(mtcars, mpg, cyl, wt)
```

A select step is relatively simple to implement since `dplyr::select()` operates on names.  A simple example of a JSON input and engine for column selection is shown below.

```json
{
  "step": "select",
  "parameters" : {
    "keep": ["mpg", "cyl", "wt"]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="select", parameters=list(keep=list("mpg", "cyl", "wt"))), 'data/myfile.json', auto_unbox = TRUE)
```

The simplest engine runs `dplyr::select()` directly on the column names:

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, tidyselect::any_of(unlist(inp$parameters$keep)))
print(data)
```

::: {.callout-note}
An alternate approach uses `rlang::syms()` as follows:
```{r}
#| eval: false
build_select <- function(x) {
  rlang::syms(x[["keep"]])
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, !!!build_select(inp$parameters))
print(data)
```

Here, `build_select()` returns a list of arguments which are spliced into `dplyr::select()` using `!!!`.  The expression

```{r}
#| eval: false
dplyr::select(mtcars, !!!build_select(inp$parameters))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::select(mtcars, mpg, cyl, wt)
```

Although it achieves the same outcome as using `tidyselect::any_of()`, the tidyselect approach is preferred since we do not need to construct expressions when using `dplyr::select()`.
:::

### The Use of `any_of()` {#sec-wrangling-anyof}

`tidyselect::any_of()` is a tidy-select helper used to convert character vectors into column selection specifications.

For example, interactive selection works with unquoted column names:

```{r}
#| eval: false
dplyr::select(mtcars, mpg, cyl, wt)
```

but when column names are stored in a vector, teh same code fails:

```{r}
#| eval: false
cols <- c("mpg", "cyl", "wt")
dplyr::select(mtcars, cols)
```

Using `any_of()` solves the problem:

```{r}
#| eval: false
cols <- c("mpg", "cyl", "wt")
dplyr::select(mtcars, tidyselect::any_of(cols))
```

`any_of()` silently ignores missing columns, does not evaluate expressions or use data masking, and returns a tidy-select specification, which is just what `select()` expects.

::: {.callout-important}
### Note
`any_of()` is different from `!!!syms()`.  The latter splices multiple symbols as expressions, which `any_of()` builds a single column selection object.  Using `!!!syms()` would fail because `select()` expects a tidy-select specification, not multiple specifications.
:::

### Dropping Columns

`dplyr::select()` can be used to drop columns.  For example, below the columns *mpg*, *cyl* and *wt* are dropped, whilst all other columns are retained.

```{r}
dplyr::select(mtcars, -mpg, -cyl, -wt)
```

The same expression can be written as

```{r}
#| eval: false
dplyr::select(mtcars, -c(mpg, cyl, wt))
```

Tidyselect recognizes the `-` operator as a negative instruction.  We can therefore construct a column removal function in the same manner as the column selection function.

```json
{
  "step": "select",
  "parameters" : {
    "drop": ["mpg", "cyl", "wt"]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="select", parameters=list(drop=list("mpg", "cyl", "wt"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, -tidyselect::any_of(unlist(inp$parameters$drop)))
print(data)
```

### Selecting and Dropping Columns

To build a more generalized approach that responds to both keep **and** drop columns we would like to build a function that returns a tidyselect object.  This object could be spliced into `dplyr::select()` for selecting and dropping columns.

If we write this function:

```{r}
build_select <- function(x) {
  list(
    if (!is.null(x[["keep"]])) tidyselect::any_of(unlist(x[["keep"]])),
    if (!is.null(x[["drop"]])) -tidyselect::any_of(unlist(x[["drop"]]))
  )
}
```

we find it works well when we keep columns:

```json
{
  "step": "select",
  "parameters" : {
    "keep": ["mpg", "cyl", "wt"]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="select", parameters=list(keep=list("mpg", "cyl", "wt"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, !!!build_select(inp$parameters))
print(data)
```

but fails when we try to drop columns:

```json
{
  "step": "select",
  "parameters" : {
    "drop": ["mpg", "cyl", "wt"]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="select", parameters=list(drop=list("mpg", "cyl", "wt"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| error: true
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, !!!build_select(inp$parameters))
```

This is because we cannot compute a negated selection and then inject it.  In order to build a function we cannot return a selection, we must return an expression:

```{r}
build_select <- function(x) {
  rtn <- list()

  if (!is.null(x[["keep"]])) {
    rtn <- c(rtn, list(
      tidyselect::any_of(unlist(x[["keep"]]))
    ))
  }
  if (!is.null(x[["drop"]])) {
    rtn <- c(rtn, list(
      rlang::expr(-tidyselect::any_of(unlist(!!x[["drop"]])))
    ))
  }
  rtn
}
```

The expression

```{r}
#| eval: false
rlang::expr(-tidyselect::any_of(unlist(!!x[["drop"]])))
```

is returned as

```{r}
#| echo: false
x <- inp$parameters
rlang::expr(-tidyselect::any_of(unlist(!!x[["drop"]])))
```

Now, dropping columns will work.

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="select", parameters=list(drop=list("mpg", "cyl", "wt"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, !!!build_select(inp$parameters))
print(data)
```

Plus, the two can be used together

```json
{
  "step": "select",
  "parameters" : {
    "keep": ["mpg", "cyl", "wt"],
    "drop": "wt"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="select", parameters=list(keep=list("mpg", "cyl", "wt"), drop="wt")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, !!!build_select(inp$parameters))
print(data)
```

## Keeping Unique Rows

### `dplyr::distinct()`

`dplyr::distinct()` is used to retain unique rows. It can be used in two manners, returning only the unique values in the specified columns or returning all rows, keeping  just the first row when there are multiple matches.

### Returning Only Selected Columns

`dplyr::distinct()` appears to work in a similar manner to `dplyr::select()`, taking one or more expressions separated by commas to specify columns.  For example, below the unique values from the columns *gear*, *carb* are returned.

```{r}
dplyr::distinct(mtcars, gear, carb)
```

It is, however, using data masking to interpret the arguments, as opposed to tidy select.  `dplyr::distinct()` **can** accept an arbitrary expression, not just column names.  Therefore we can write an expression like:

```{r}
dplyr::distinct(mtcars, gear + carb)
```

Since we use data masking and not tidy select, the following is not allowed:

```{r}
#| error: true
cols <- c("gear", "carb")
dplyr::distinct(dplyr::any_of(cols))
```

Therefore, in order to build a `distinct()` step programmatically, we need to evaluate a data-masked expression.  A simple example of a JSON input and engine for this approach is shown below.

```json
{
  "step": "distinct",
  "parameters": {
    "cols": ["gear", "carb"]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="distinct", parameters=list(cols=list("carb", "gear"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_distinct <- function(x) {
  rlang::syms(x$cols)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::distinct(mtcars, !!!build_distinct(inp$parameters))
print(data)
```

Using symbols for column names represents the simplest possible form of expression.  Although they resemble column selections, these arguments are evaluated as data-masked expressions rather than through tidy select.

### Returning All Columns

`dplyr::distinct()` identifies rows with unique values.  If more than one row is satisfied,it will return the first match.  It can return either selected columns, as demonstrated above, or, if argument `.keep_all=TRUE`, will return all columns.

```{r}
dplyr::distinct(mtcars, gear, carb, .keep_all = TRUE)
```

Our engine can easily be extended to account for the use of `.keep_all` as follows.

```json
{
  "step": "distinct",
  "parameters": {
    "cols": ["gear", "carb"],
    "keep_all": true
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="distinct", parameters=list(cols=list("carb", "gear"), keep_all=TRUE)), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_distinct <- function(x) {
  rlang::syms(x$cols)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::distinct(mtcars, !!!build_distinct(inp$parameters), .keep_all = isTRUE(inp$parameters$keep_all))
print(data)
```

The change in code occurs at `dplyr::distinct()`, and does not affect the way expressions are built.  The use of `isTRUE` accounts for missing values (`isTRUE(NULL)` equates to `FALSE`).

### Expression Handling

Since `distinct()` relies on data masking, we can replace the broken-down approach with expression parsing.  @sec-expression-parser introduced a method for safe expression handling.  This method can be applied to `distinct()`.

```{r}
allowed_operators <- c("+", "-", "*", "/", "^", 
                       ">", ">=", "<", "<=", "==", "!=", 
                       "&&", "||", "!", "c", "mean", "sd",
                       "&", "|")
                       
validate_expr <- function(expr) {
  
  # Atomic values (numbers, strings, logicals) are always safe
  if (rlang::is_atomic(expr)) {
    return(TRUE)
  }

  # Symbols (variables and column names) are safe
  if (rlang::is_symbol(expr)) {
    return(TRUE)
  }

  # Check function calls
  if (rlang::is_call(expr)) {
    fn <- rlang::call_name(expr)

    # Parentheses are grouping operators, not function calls
    if (fn == "(") {
      validate_expr(expr[[2]])
      return(TRUE)
    }

    # Reject disallowed operators
    if (!fn %in% allowed_operators) {
      stop("Disqualified operator: ", fn, call. = FALSE)
    }

    # Recursively validate arguments
    for (i in 2:length(expr)) {
      validate_expr(expr[[i]])
    }

    return(TRUE)
  }

  # Anything else is rejected
  stop("Unsupported expression type", call. = FALSE)
}

build_distinct_expr <- function(x) {
  expr <- rlang::parse_expr(x)
  validate_expr(expr)
  expr
}
```

Here's a simple example to illustrate returning unique rows using an expression

```json
{
  "step": "distinct",
  "parameters": {
    "expression": "gear + carb"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="distinct", parameters=list(expression="gear + carb")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
dplyr::distinct(mtcars, !!build_distinct_expr(inp$parameters$expression))
```

## Renaming Columns

### `dplyr::rename()`

`dplyr::rename()` changes the names of existing columns using `new_name = old_name` syntax.  Rename uses tidy-select format for arguments, taking a named list.

### Renaming a Single Column

`dplyr::rename()` works on names, in a similar manner to `dplyr::select()`.  Renaming a single column is demonstrated below.

```{r}
dplyr::rename(mtcars, miles_per_gallon = mpg)
```

In JSON format this can be represented as:

```json
{
  "step": "rename",
  "parameters": {
    "old_name": "mpg",
    "new_name": "miles_per_gallon"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="rename", parameters=list(old_name="mpg", new_name="miles_per_gallon")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_rename <- function(x) {
  setNames(
    list(x$old_name), 
    x$new_name
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::rename(mtcars, !!!build_rename(inp$parameters))
print(data)
```

Here we cannot use `tidyselect::any_of()` as it returns a column selection, not a rename mapping.  In order to programmatically build a rename step we must splice the named arguments using `!!!`.  The named arguments are built using `setNames()`.

### Renaming Multiple Columns

The same concept can be used when extending to multiple columns.

```{r}
dplyr::rename(mtcars, miles_per_gallon = mpg, cylinders = cyl, weight = wt)
```

```json
{
  "step": "rename",
  "parameters": {
    "mpg": "miles_per_gallon",
    "cyl": "cylinders",
    "wt": "weight"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="rename", parameters=list(mpg="miles_per_gallon", cyl="cylinders", wt="weight")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_rename <- function(x) {
  setNames(
    as.list(names(x)),
    unname(x)
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::rename(mtcars, !!!build_rename(inp$parameters))
print(data)
```

*build_rename(inp$rename)* returns:

```{r}
build_rename(inp$parameters)
```

and 

```{r}
#| eval: false
dplyr::rename(mtcars, !!!build_rename(inp$parameters))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::rename(mtcars, miles_per_gallon = mpg, cylinders = cyl, weight = wt)
```

An alternative, more verbos approach to multiple column renaming is:

```json
{
  "step": "rename",
  "parameters": {
    "rename_mapping": [
      {"old_name": "mpg", "new_name": "miles_per_gallon"},
      {"old_name": "cyl", "new_name": "cylinders"},
      {"old_name": "wt", "new_name": "weight"}
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="rename", parameters=list(rename_mapping=list(list(old_name="mpg", new_name="miles_per_gallon"), list(old_name="cyl", new_name="cylinders"), list(old_name="wt", new_name="weight")))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_rename <- function(x) {
  setNames(
    lapply(x$rename_mapping, function(y) y$old_name),
    sapply(x$rename_mapping, function(y) y$new_name)
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::rename(mtcars, !!!build_rename(inp$parameters))
print(data)
```

Although `rename()` uses tidy-select, it does not accept selections directly.  Instead tidy-select expressions must appear on the righthand side of named arguments the define the new column names.

## Ordering Rows

### `dplyr::arrange()`

`dplyr::arrange()` orders rows of a data frame by the values of selected columns.  `dplyr::arrange()` can sort rows using ascending or descending order and generally ignores grouping.

### Order on a Single Column

Ordering (ascending) on a single column is quite straightforward to implement.

```{r}
dplyr::arrange(mtcars, carb)
```

The JSON structure and engine look as follows:

```json
{
  "step": "arrange",
  "parameters": {
    "col": "carb"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="arrange", parameters=list(col="carb")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_arrange <- function(x) {
  rlang::sym(x$col)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::arrange(mtcars, !!build_arrange(inp$parameters))
print(data)
```

`dplyr::arrange()` takes arguments as data-masked.  Passing a tidy-select construct does not lead to an error the columns are not sorted either.  This version will not error but the data frame will be returned unchanged:

```{r}
#| message: false
build_arrange <- function(x) {
  x$col
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::arrange(mtcars, !!build_arrange(inp$parameters))
print(data)
```

### Order on Multiple Columns

The approach can easily be extended to account for ordering (ascending) on multiple columns.

```{r}
dplyr::arrange(mtcars, carb, mpg)
```

The JSON structure and engine look as follows:

```json
{
  "step": "arrange",
  "parameters": {
    "cols": ["carb", "mpg"]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="arrange", parameters=list(cols=list("carb", "mpg"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_arrange <- function(x) {
  rlang::syms(x$cols)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::arrange(mtcars, !!!build_arrange(inp$parameters))
print(data)
```

### Ascending and Descending Orders

Building a generic arrange function to account for multiple ascending and descending orders takes a little thought.  We cannot apply more than one `dplyr::arrange()` step, as later steps would override earlier ones leading to the data frame being re-sorted each time.  `dplyr::arrange()` is executed with no knowledge of previous sorting.  The entire `dplyr::arrange()` expression must be built into a single expression.  The example below loses any ordering from **carb** once it is ordered by **mpg**.

```{r}
dplyr::arrange(mtcars, carb) |>
  dplyr::arrange(mpg)
```

The `dplyr::arrange()` expression can be built up within an `lapply()` loop, accounting for ascending or descending order.  For example, the following sorts by gear (ascending), followed by cyl (descending) and mpg (ascending)

```{r}
dplyr::arrange(mtcars, gear, dplyr::desc(cyl), mpg)
```

A JSON format and engine could look like:

```json
{
  "step": "arrange",
  "parameters": {
    "vars" : [
      {
        "column": "gear",
        "direction": "asc"
      },
      {
        "column": "cyl",
        "direction": "desc"
      },
      {
        "column": "mpg"
      }
    ]
  }
}

```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="arrange", parameters=list(vars=list(list(column="gear", direction="asc"), list(column="cyl", direction="desc"), list(column="mpg")))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_arrange <- function(x) {
  lapply(x$vars, function(y) {
    if (isTRUE(y[['direction']] == "desc")) {
      rlang::call2("desc", rlang::sym(y[['column']]), .ns = "dplyr")
    } else { 
      rlang::sym(y[['column']])
    }
  })
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::arrange(mtcars, !!!build_arrange(inp$parameters))
print(data)
```

`build_arrange()` takes a list for ordering.  Each element is a list containing a column name and optional direction for sorting with the default being *ascending*.  When sorted as *ascending** `sym(column)` is returned and when sorted as *descending* `desc(sym(column))` is returned.  The use of `.ns` in `rlang::call2()` ensures that the namespace `dplyr` is prepended to `desc()`.  `build_arrange(inp$arrange)` returns:

```{r}
build_arrange(inp$parameters)
```

and 

```{r}
#| eval: false
dplyr::arrange(mtcars, !!!build_arrange(inp$parameters))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::arrange(mtcars, gear, dplyr::desc(cyl), mpg)
```

### Expression Handling

Since `arrange()` relies on data masking, like `distinct()`, we can replace the broken-down approach with expression parsing.  The same safe expression handling code from @sec-expression-parser can be applied to `arrange()`.

```{r}
allowed_operators <- c("desc")

validate_expr <- function(expr) {
  
  # Atomic values (numbers, strings, logicals) are always safe
  if (rlang::is_atomic(expr)) {
    return(TRUE)
  }

  # Symbols (variables and column names) are safe
  if (rlang::is_symbol(expr)) {
    return(TRUE)
  }

  # Check function calls
  if (rlang::is_call(expr)) {
    fn <- rlang::call_name(expr)

    # Parentheses are grouping operators, not function calls
    if (fn == "(") {
      validate_expr(expr[[2]])
      return(TRUE)
    }

    # Reject disallowed operators
    if (!fn %in% allowed_operators) {
      stop("Disqualified operator: ", fn, call. = FALSE)
    }

    # Recursively validate arguments
    for (i in 2:length(expr)) {
      validate_expr(expr[[i]])
    }

    return(TRUE)
  }

  # Anything else is rejected
  stop("Unsupported expression type", call. = FALSE)
}

build_arrange_expr <- function(x) {
  lapply(x, function(y) {
    if (startsWith(y, "desc")) y <- paste0("dplyr::", y)
    expr <- rlang::parse_expr(y)
    validate_expr(expr)
    expr
  })
}
```

Here's an example to illustrate sorting using an expression with a mixture of ascending and descending orders.

```json
{
  "step": "arrange",
  "parameters": {
    "vars": ["gear", "desc(cyl)", "mpg"]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(step="arrange", parameters=list(vars=list("gear", "desc(cyl)", "mpg"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
dplyr::arrange(mtcars, !!!build_arrange_expr(inp$parameters$vars))
```
