# Combining Tables

The dplyr library has six functions used to combine data frames on common variables.  Four of these are mutating joins (combining columns) and two are filtering joins (filtering rows).

Assuming we have two data frames, `x` and `y`:

- Mutating Joins
    - **inner_join(`x`, `y`)** Return all rows from both `x` and `y` in which there are matching rows in both data frames.  Rows that do not match in both data frames are dropped.
    - **left_join(`x`, `y`)** Return all rows from `x` and any matching rows from `y`.  Rows in `x` without a match in `y` will contain `NA` values in the new columns from `y`.  Rows in `y` without a match in `x` are dropped.
    - **right_join(`x`, `y`)** Return all rows from `y` and any matching rows from `x`.  Rows in `y` without a match in `x` will contain `NA` values in the new columns from `x`.  Rows in `x` without a match in `y` are dropped.
    - **full_join(`x`, `y`)** Return all rows and columns from both `x` and `y`.  Where there are no matching values `NA` will be inserted.
- Filtering Joins
    - **semi_join(`x`, `y`)** Return all rows from `x` that have a match in `y`.  Similar to `inner_join()` but only keeps columns from `x`.
    - **anti_join(`x`, `y`)** Return all rows from `x` that do **not** have a match in `y`.  This is often used to find rows in `x` with no corresponding data in `y`.

## Building Joins

### `dplyr::xxx_join()`

Each join in {dplyr} requires two data frames (`x` and `y`) along with one or more columns that specify the keys (`by`).  If `by` is missing, then all common column names are used.

```{r}
#| eval: false
## join x and y using matches in `col1`
dplyr::inner_join(x, y, by = "col1")

## join x and y using matches in both `col1` and `col21
dplyr::inner_join(x, y, by = c("col1", "col2"))
```

If the column names in `x` and `y` differ, then a named vector can be used.  For example, if column names in `x` are `col1`, `col2` and `col3`, and column names in `y` are `A`, `B` and `C`, `x` and `y` could be joined as

```{r}
#| eval: false
## join x and y using matches in `col1` and `col2`
dplyr::inner_join(x, y, by = c(col1 = "A", col2 = "B"))
```

In order to develop the concept we'll need two datasets.

```{r}
customers <- dplyr::tibble(
  customer_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Carol", "Dan")
)

orders <- dplyr::tibble(
  customer_id = c(1, 1, 3, 5),
  order_id = c(101, 102, 103, 104),
  amount = c(50, 75, 100, 20)
)
```

```{r}
customers
orders
```

```{r}
dplyr::inner_join(customers, orders, by = "customer_id")
```
Inner join keeps only the matching keys.

```{r}
dplyr::left_join(customers, orders, by = "customer_id")
```
Left join keeps all customers.

```{r}
dplyr::right_join(customers, orders, by = "customer_id")
```
Right join keeps all orders.

```{r}
dplyr::full_join(customers, orders, by = "customer_id")
```
Full join keeps everything.

```{r}
dplyr::semi_join(customers, orders, by = "customer_id")
```
Semi join keeps customers with at least one order but no columns from `orders` are added and there is no duplication for multiple matches.

```{r}
dplyr::anti_join(customers, orders, by = "customer_id")
```
Anti join keeps customers without any orders.

### A Universal Join Engine

A single engine can be used to join tables.  It works by identifying the type of join and executing the relevant function.  Here's a typical JSON input:

```json
{
    "join": {
        "type": "left",
        "table_1": "customers",
        "table_2": "orders",
        "by" : {
            "customer_id": "customer_id"
        }
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(join=list(type="left", table_1="customers", table_2="orders", by=list(customer_id="customer_id"))), 'data/myfile.json', auto_unbox = TRUE)
```

and here's an engine, assuming our data frames are stored in a named list, tables, such that:

```{r}
tables <- list(
    customers = customers,
    orders = orders
)
```

```{r}
join_operators <- list(
    inner = dplyr::inner_join,
    left = dplyr::left_join,
    right = dplyr::right_join,
    full = dplyr::full_join,
    semi = dplyr::semi_join,
    anti = dplyr::anti_join
)

build_join <- function(type, tab1, tab2, by) {
    join_fn <- join_operators[[type]]
    join_fn(tab1, tab2, by)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

join <- inp$join
data <- build_join(
    join$type, 
    tables[[join$table_1]],
    tables[[join$table_2]],
    unlist(join$by)
)
print(data)
```

`join_operators` is a function lookup table which maps a join type (eg, *inner*, *left*, ...) to a dplyr function (eg `dplyr::inner_join()`, `dplyr::left_join()`).

The same engine can be used to process any of the joins.  For example:

```json
{
    "join": {
        "type": "anti",
        "table_1": "customers",
        "table_2": "orders",
        "by" : {
            "customer_id": "customer_id"
        }
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(join=list(type="anti", table_1="customers", table_2="orders", by=list(customer_id="customer_id"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- build_join(
    join$type, 
    tables[[join$table_1]],
    tables[[join$table_2]],
    unlist(join$by)
)
print(data)
```

### Multiple Joins

Multiple joins can be composed together.  Let's introduce an additional table:

```{r}
payments <- dplyr::tibble(
  order_id = c(101, 102, 103, 104),
  paid = c(TRUE, TRUE, FALSE, TRUE),
  date = c("2002-10-20", "2003-05-06", NA, "2002-08-19")
)

tables[["payments"]] <- payments
```

```json
{
    "joins": [
        {
            "type": "left",
            "table_1": "customers",
            "table_2": "orders",
            "by" : {
                "customer_id": "customer_id"
            }
        },
        {
            "type": "left",
            "table_2": "payments",
            "by" : {
                "order_id": "order_id"
            }
        }
    ]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(joins=list(list(type="left", table_1="customers", table_2="orders", by=list(customer_id="customer_id")), list(type="left", table_2="payments", by=list(order_id="order_id")))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- data.frame()
for (j in inp$joins) {
    if (is.null(j[["table_1"]])) {
        tab1 <- data
    } else {
        tab1 <- tables[[j[["table_1"]]]]
    }
    data <- build_join(
        j[["type"]],
        tab1,
        tables[[j[["table_2"]]]],
        unlist(j[["by"]])
    )
}

print(data)
```

In this approach, the first join step defines the starting table and subsequent joins operate on the current result.
