# Transformation

## Filtering Rows

### `dplyr::filter()`

`dplyr::filter()` retains rows of a data frame that satisfy one or more logical conditions.

### Filtering a Single Value {#sec-filter-single-value}

Starting with a simple example, here we return all rows in which `mpg > 30`.

```{r}
dplyr::filter(mtcars, mpg > 30)
```

The filter itself can be split into a column identifier, an operator and a value.  A JSON input for such as filter could be:

```json
{
  "filter": {
    "col": "mpg",
    "operator": ">",
    "value": 30
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(col="mpg", operator=">", value=30)), 'data/myfile.json', auto_unbox = TRUE)
```


In R, operators such as `==`, `>`, `<`, etc are functions.

```{r}
`==`(3, 4)
`==`(4, 4)
```

We can therefore build up a logical expression for a filter using operators as functions and our engine could be:

```{r}
build_filter <- function(x) {
  rlang::call2(x$operator, rlang::sym(x$col), x$value)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- dplyr::filter(mtcars, !!build_filter(inp$filter))
print(data)
```

Here, we build an expression using the name of the column (`mpg`).  *build_filter(inp$filter)* returns:

```{r}
build_filter(inp$filter)
```

and

```{r}
#| eval: false
dplyr::filter(mtcars, !!build_filter(inp$filter))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::filter(mtcars, mpg > 30)
```

::: {.callout-tip}
The filter step can be achieved using functions other than `rlang::call2()`, such as the base R functions `get()`, `do.call()`,`match.fun()` and `call()`, but `rlang::call2()` is a preferred way of working as it is tidy-eval compatible.  This means that we can define an expression and use it later inside a dplyr verb (eg `dplyr::filter(!!expr)`).  The expression is evaluated the time of execution, not definition.  `call()` works in a similar manner but is not as flexible and `get()`, `do.call()` and `match.fun()` execute immediately.
:::

### Filtering Against a List of Values {#sec-filter-list-values}

It is relatively simple to extend this logic to work for multiple values.  In the example below, `value` has been replaced by a list of values and the operator is called `in`.

```{r}
dplyr::filter(mtcars, cyl %in% c(4, 5, 6))
```

```json
{
  "filter": {
    "col": "cyl",
    "operator": "in",
    "value": [4, 5, 6]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(col="cyl", operator="in", value=list(4,5,6))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
build_filter <- function(x) {
  if (x$operator == "in") {
    rlang::call2("%in%", rlang::sym(x$col), unlist(x$value))
  }
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- dplyr::filter(mtcars, !!build_filter(inp$filter))
print(data)
```

### Mapping Operators {#sec-filter-mapping-operators}

Building a 1:1 mapping from operator names to functions provides two advantages.  Firstly, the operator can be checked to confirm that it is allowed and, secondly, we can use this method to use a new name for an existing function.  Our operator mapping is a named vactor as follows:

```{r}
operators <- c(
  "==" = "==",
  ">" = ">",
  ">=" = ">=",
  "<" = "<",
  "<=" = "<=",
  "!=" = "!=",
  "in" = "%in%"
)
```

Here we can see that most operators map directly to their name with the exception of `%in%` which is referred to as `in`.  This allows us to combine all these operators into a single function:

```{r}
build_filter <- function(x) {
  if (x$operator %in% names(operators)) {
    rlang::call2(operators[x$operator], rlang::sym(x$col), unlist(x$value))
  }
}
```

The examples discussed in @sec-filter-single-value and @sec-filter-list-values can be run using the same function:

```json
{
  "filter": {
    "col": "mpg",
    "operator": ">",
    "value": 30
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(col="mpg", operator=">", value=30)), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- dplyr::filter(mtcars, !!build_filter(inp$filter))
print(data)
```



```json
{
  "filter": {
    "col": "cyl",
    "operator": "in",
    "value": [4, 5, 6]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(col="mpg", operator=">", value=30)), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- dplyr::filter(mtcars, !!build_filter(inp$filter))
print(data)
```

### Multiple Operators

Filter steps can be combined using `AND` and `OR` logic to build more complex cases.

#### Combining Filters using `AND`

`AND` is the default behavior when multiple filter conditions are supplied.  Rows are returned if they meet all of the conditions.  The example below filters on `carb == 2` and `mpg > 30`.

```{r}
dplyr::filter(mtcars, carb == 2, mpg > 30)
```

This can also be represented as

```{r}
#| eval: false
dplyr::filter(mtcars, carb == 2 & mpg > 30)
```

or 

```{r}
#| eval: false
dplyr::filter(mtcars, carb == 2) |>
  dplyr::filter(mtcars, mpg > 30)
```


```json
{
  "filter" : [
    {
      "col": "carb",
      "operator": "==",
      "value": 2
    },
    {
      "col": "mpg",
      "operator": ">",
      "value": 30
    }
  ]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(list(col="carb",operator="==",value=2), list(col="mpg",operator=">",value=30))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
operators <- c(
  "==" = "==",
  ">" = ">",
  ">=" = ">=",
  "<" = "<",
  "<=" = "<=",
  "!=" = "!=",
  "in" = "%in%"
)

build_filter <- function(x) {
  if (x$operator %in% names(operators)) {
    rlang::call2(operators[x$operator], rlang::sym(x$col), unlist(x$value))
  }
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- mtcars
for (filter_step in inp$filter) {
  data <- dplyr::filter(data, !!build_filter(filter_step))
}
print(data)
```

Here we apply each filter in a stepwise manner, passing the output from each to the next.  This combines all filters using `AND` logic.

#### Combining Filters using `OR`

`AND` works naturally with repeated calls to `filter()` because each call reduces the data.  `OR` cannot be implemented stepwise because each call would discard rows that might be needed by later conditions.  One way to work with `OR` logic is to build a single expression through concatenating conditions with the `|` operator and evaluating the resulting combination.  

```{r}
dplyr::filter(mtcars, carb == 2 | mpg > 30)
```

```json
{
  "filter" : [
    {
      "col": "carb",
      "operator": "==",
      "value": 2
    },
    {
      "col": "mpg",
      "operator": ">",
      "value": 30
    }
  ]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(list(col="carb",operator="==",value=2), list(col="mpg",operator=">",value=30))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
operators <- c(
  "==" = "==",
  ">" = ">",
  ">=" = ">=",
  "<" = "<",
  "<=" = "<=",
  "!=" = "!=",
  "in" = "%in%"
)

build_filter <- function(x) {
  if (x$operator %in% names(operators)) {
    rlang::call2(operators[x$operator], rlang::sym(x$col), unlist(x$value))
  }
}

combine_filters_or <- function(x) {
  ## build each condition
  filter_steps <- lapply(x, function(y) {
    build_filter(y)
  })

  ## combine conditions with 'OR'
  Reduce(
    f = function(lhs, rhs) rlang::call2("|", lhs, rhs),
    x = filter_steps
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

data <- dplyr::filter(mtcars, !!combine_filters_or(inp$filter))
print(data)
```

Here we build a separate expression for each filter and then combine them using `Reduce()` to generate a single `OR` expression.  The individual expressions are:

```{r}
x <- inp$filter
filter_steps <- lapply(x, function(y) {
  build_filter(y)
})
print(filter_steps)
```

and combined:

```{r}
Reduce(
  f = function(lhs, rhs) rlang::call2("|", lhs, rhs),
  x = filter_steps
)
```

`Reduce()` combines a sequence of values into a single result by repeatedly applying a binary function, in this case the `OR` function, `|`.

`purrr::reduce()` may be used in place of `Reduce()`.  It differs in order of arguments (data before function to accommodate piping).

```{r}
purrr::reduce(filter_steps, \(lhs, rhs) rlang::call2("|", lhs, rhs))
```

::: {.callout-note}
Care should be taken when working with `Reduce()` and non-associative logic.  For example:

```{r}
my_vals <- c(10, 5, 2)
# by default, Reduce() proceeds from left to right
# (10 - 5) - 2 = 3
Reduce("-", my_vals)

# RIGHT=TRUE forces Reduce() to proceed from right to left
# 10 - (5 - 2) = 7
Reduce("-", my_vals, right = TRUE)
```
:::

### Combining `AND` and `OR` {#sec-filter-and-or}

As noted above, there is a key distinction between `AND` and `OR` logic.  `AND` is sequentially reducible (output is TRUE only if all conditions are satisfied) whereas `OR` must be expressed as a single logical expression.  We can build a function that combines `AND` and `OR` logic but must account for the use of parentheses.  For example, since `&` takes precedence over `|`, the following are not equivalent.

```{r}
dplyr::filter(mtcars, carb == 2 | mpg > 30 & cyl == 4)
```

```{r}
dplyr::filter(mtcars, (carb == 2 | mpg > 30) & cyl == 4)
```

The first,

```{r}
#| eval: false
dplyr::filter(mtcars, carb == 2 | mpg > 30 & cyl == 4)
```

is equivalent to

```{r}
#| eval: false
dplyr::filter(mtcars, carb == 2 | (mpg > 30 & cyl == 4))
```

::: {.callout-note}
R does not store parentheses, it stores the expression structure.  The use of parentheses changes the structure of the expression tree.  More information on building expressions and the abstract syntax trees can be found in *Advanced R*.[^1]
:::

[^1]: Wickham, H. (2019). Advanced R (2nd ed.). Chapman & Hall/Routledge [https://adv-r.hadley.nz/expressions.html#ast](https://adv-r.hadley.nz/expressions.html#ast)

We can use a nested JSON structure to build our instruction set which will allow for the use of parentheses and ensure that this example works correctly:

```{r}
dplyr::filter(mtcars, (carb == 2 | mpg > 30) & cyl == 4)
```

```json
{
  "filter": {
    "logic": "AND",
    "steps": [
      {
        "logic": "OR",
        "steps": [
          {"col": "carb", "operator": "==", "value": 2},
          {"col": "mpg", "operator": ">", "value": 30}
        ]
      },
      {"col": "cyl", "operator": "==", "value": 4}
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(logic="AND", steps=list(list(logic="OR", steps=list(list(col="carb",operator="==",value=2), list(col="mpg",operator=">",value=30))), list(col="cyl",operator="==",value=4)))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
operators <- c(
  "==" = "==",
  ">" = ">",
  ">=" = ">=",
  "<" = "<",
  "<=" = "<=",
  "!=" = "!=",
  "in" = "%in%"
) 
logic_operators <- c(
  "AND" = "&",
  "OR" = "|"
)

build_filter <- function(x) {
  if (is.null(x$logic)) {
    ## single condition - no logic
    rlang::call2(operators[x$operator], rlang::sym(x$col), x$value)
  } else if (!is.null(x$steps)) {
    ## combine sub-filters through recursion
    filter_steps <- lapply(x$steps, build_filter)
    op <- logic_operators[x$logic]
    Reduce(
      function(lhs, rhs) rlang::call2(op, lhs, rhs),
      filter_steps
    ) 
  } else {
    stop("Invalid JSON")
  }
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_filter(inp$filter)
data <- dplyr::filter(mtcars, !!expr)
print(data)
```

In the fuction `build_filter()`, the input list is checked for the presence of an element with the name *logic*.  If none exists then there must be a single filter condition and an expression for the condition is returned.  If *logic* is present then filters are combined using `Reduce()`, running the `build_filter()` function on all subsequent filter steps recursively.  The hierarchy built into the JSON carries through to the encoding in the expression tree.  The individual expressions can be viewed as:

```{r}
build_individual_expressions <- function(x) {
  if (is.null(x$logic)) {
    ## single condition - no logic
    rlang::call2(operators[x$operator], rlang::sym(x$col), x$value)
  } else if (!is.null(x$steps)) {
    ## run function on all children
    lapply(x$steps, build_individual_expressions)
  }
}

individual_expressions <- build_individual_expressions(inp$filter)
print(individual_expressions)
```

and the final expression is

```{r}
expr <- build_filter(inp$filter)
print(expr)
```

Changing the JSON hierarchy leads to a different expression tree and hence different result.

```json
{
  "filter": {
    "logic": "OR",
    "steps": [
      {"col": "carb", "operator": "==", "value": 2},
      {
        "logic": "AND",
        "steps": [
          {"col": "mpg", "operator": ">", "value": 30},
          {"col": "cyl", "operator": "==", "value": 4}
        ]
      }
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(logic="OR", steps=list(list(col="carb",operator="==",value=2), list(logic="AND", steps=list(list(col="mpg",operator=">",value=30), list(col="cyl",operator="==",value=4)))))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_filter(inp$filter)
data <- dplyr::filter(mtcars, !!expr)
print(data)
```

### Negating with `NOT`

This approach naturally extends to the use of `NOT` to negate a condition.  `NOT` is a unary operator (it takes a single expression), whereas `AND` and `OR` are logical combiners (they take more than one expression).  Therefore `NOT` has to be treated differently.

```{r}
dplyr::filter(mtcars, !(carb == 2 | mpg > 30) & cyl == 4)
```


```json
{
  "filter": {
    "logic": "AND",
    "steps": [
      {
        "logic": "NOT",
        "steps": {
          "logic": "OR",
          "steps": [
            {"col": "carb", "operator": "==", "value": 2},
            {"col": "mpg", "operator": ">", "value": 30}
          ]
        }
      },
      {"col": "cyl", "operator": "==", "value": 4}
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(logic="AND", steps=list(list(logic="NOT", steps=list(logic="OR", steps=list(list(col="carb",operator="==",value=2), list(col="mpg",operator=">",value=30)))), list(col="cyl",operator="==",value=4)))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
operators <- c(
  "==" = "==",
  ">" = ">",
  ">=" = ">=",
  "<" = "<",
  "<=" = "<=",
  "!=" = "!=",
  "in" = "%in%"
) 
logic_operators <- c(
  "AND" = "&",
  "OR" = "|"
)

build_filter <- function(x) {
  if (is.null(x$logic)) {
    ## single condition - no logic
    return(rlang::call2(operators[x$operator], rlang::sym(x$col), x$value))
  }
  
  if (x$logic == "NOT") {
    inner_expression <- build_filter(x$steps)
    return(rlang::call2("!", inner_expression))
  }
  
  if (x$logic %in% names(logic_operators)) {
    ## combine sub-filters through recursion
    filter_steps <- lapply(x$steps, build_filter)
    op <- logic_operators[x$logic]
    return(
      Reduce(
        function(lhs, rhs) rlang::call2(op, lhs, rhs),
        filter_steps
      ) 
    )
  } else {
    stop("Invalid JSON")
  }
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_filter(inp$filter)
data <- dplyr::filter(mtcars, !!expr)
print(data)
```

We now have three outputs from `build_filter()`:

- if there is no *logic* then it must be a single condition and therefore the expression is returned.
- if *logic* = `NOT` then run `build_filter()` on steps under `NOT` and return the negated expression.
- if *logic* = `AND` or `OR` then run `build_filter()` on steps under the condition and return the combined expression.

The final expression passed to `dpyr::filter()` is:

```{r}
expr
```

### The Challenge of Values and Column Names {#sec-combining-values-and-column-names}

It is important to note the difference between filtering against a value vs. filtering against another column.  For example, in R each of the following are valid:


```{r}
dplyr::filter(mtcars, wt > drat)
dplyr::filter(mtcars, wt > 4)
```

However, when we use an engine to parse code built from an instruction set we might wish to programmatically build expressions using tidy evaluation.  Unfortunately we cannot use a single approach for both column names and scalar values as `!!sym(column)` is valid but `!!sym(value)` is not.  `sym()` must only be used for symbols and scalar values must be injected directly into the expression.

```{r}
#| error: true
## This code works
x1 <- 'wt'
x2 <- 'drat'
expr <- rlang::call2(">", rlang::sym(x1), rlang::sym(x2))
dplyr::filter(mtcars, !!expr)
```
```{r}
#| error: true
## This code fails
x1 <- 'wt'
x2 <- 4
expr <- rlang::call2(">", rlang::sym(x1), rlang::sym(x2))
```

```{r}
#| error: true
## This code works
x1 <- 'wt'
x2 <- 4
expr <- rlang::call2(">", rlang::sym(x1), x2)
dplyr::filter(mtcars, !!expr)
```

A solution is to check if `value` is a column name and respond accordingly.  A simple example is shown below.  Bear in mind that the logic may need to be a little more complex than shown here.

```json
{
  "filter": {
    "logic": "AND",
    "steps": [
      {"col": "cyl", "operator": ">", "value": "drat"},
      {"col": "wt", "operator": ">", "value": 4}
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(logic="AND", steps=list(list(col="cyl", operator=">", value="drat"), list(col="wt", operator=">", value=4)))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
operators <- c(
  "==" = "==",
  ">" = ">",
  ">=" = ">=",
  "<" = "<",
  "<=" = "<=",
  "!=" = "!=",
  "in" = "%in%"
) 
logic_operators <- c(
  "AND" = "&",
  "OR" = "|"
)

build_filter <- function(x, col_names) {
  if (is.null(x$logic)) {
    ## single condition - no logic
    if (x$value %in% col_names) {
      value <- rlang::sym(x$value)
    } else {
      value <- x$value
    }
    rlang::call2(operators[x$operator], rlang::sym(x$col), value)
  } else if (!is.null(x$steps)) {
    ## combine sub-filters through recursion
    filter_steps <- lapply(x$steps, build_filter, col_names)
    op <- logic_operators[x$logic]
    Reduce(
      function(lhs, rhs) rlang::call2(op, lhs, rhs),
      filter_steps
    ) 
  } else {
    stop("Invalid JSON")
  }
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_filter(inp$filter, names(mtcars))
data <- dplyr::filter(mtcars, !!expr)
print(data)
```

The logic included to determine if a value is a column name or scalar is fragile and may fail (consider the case where `"4"` is a column name).  A better approach would be to make the intent specific.

```json
{
  "filter": {
    "logic": "AND",
    "steps": [
      {
        "col": "cyl", 
        "operator": ">", 
        "value": {
          "type": "column",
          "val": "drat"
        }
      },
      {
        "col": "wt", 
        "operator": ">", 
        "value": {
          "type": "scalar",
          "val": 4
        }
      }
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter=list(logic="AND", steps=list(list(col="cyl", operator=">", value=list(type="column", val="drat")), list(col="wt", operator=">", value=list(type="scalar", val=4))))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
operators <- c(
  "==" = "==",
  ">" = ">",
  ">=" = ">=",
  "<" = "<",
  "<=" = "<=",
  "!=" = "!=",
  "in" = "%in%"
) 
logic_operators <- c(
  "AND" = "&",
  "OR" = "|"
)

build_value <- function(v) {
  if (v$type == "column") {
    value <- rlang::sym(v$val)
  } else {
    value <- v$val
  }
}

build_filter <- function(x, col_names) {
  if (is.null(x$logic)) {
    ## single condition - no logic
    value <- build_value(x$value)
    rlang::call2(operators[x$operator], rlang::sym(x$col), value)
  } else if (!is.null(x$steps)) {
    ## combine sub-filters through recursion
    filter_steps <- lapply(x$steps, build_filter, col_names)
    op <- logic_operators[x$logic]
    Reduce(
      function(lhs, rhs) rlang::call2(op, lhs, rhs),
      filter_steps
    ) 
  } else {
    stop("Invalid JSON")
  }
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_filter(inp$filter, names(mtcars))
data <- dplyr::filter(mtcars, !!expr)
print(data)
```

## Adding and Changing Columns

### `dplyr::mutate()`

`dplyr::mutate()` is used to create new columns or modify existing ones.  

### Adding a New Column with a Constant Value

This simple case illustrates how a column, holding a single constant value, may be added using a dplyr `mutate()` function.

```{r}
dplyr::mutate(mtcars, new_col = 3)
```

The JSON implementation can be something like:

```json
{
  "mutate": {
    "new_col_name": "new_col",
    "value": 3
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(mutate=list(new_col_name="new_col", "value"=3)), 'data/myfile.json', auto_unbox = TRUE)
```

When running programmatically, the name value operator, `:=` is used in place of `=`.  Since this is a simple `mutate()` step we can build the lefthand and righthand sides directly into the `mutate()` function.

```{r}
library(rlang)

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::mutate(mtcars, !!inp$mutate$new_col_name := inp$mutate$value)
print(data)
```

When building a more generic `mutate()` step, one would think that we could use a similar syntax to `filter()`, ie construct an expression using an lefthand side, a righthand side and an operator, where the operator is `:=`.  Unfortunately, this code fails:

```{r}
#| error: true
build_mutate <- function(x) {
  rlang::call2(":=", rlang::sym(x$new_col_name), x$value)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_mutate(inp$mutate)
data <- dplyr::mutate(mtcars, !!expr)
```

The error arises because, although `:=` looks like an function, it is not.  It is a syntactic marker that only exists inside *tidy-eval* expressions.

This means that to build mutate steps programmatically we need to find an alternate approach.

#### Working with `:=`

We can still work using `:=` as long as we use it within the `mutate()` function itself.  Since `:=` can take an expression on both the lefthand and righthand sides, the following is valid:

```{r}
col <- "new_col"
val <- 3
rhs <- rlang::expr(val)
## rhs is an expression to illustrate how this method can be utilized for more complex cases
dplyr::mutate(mtcars, !!col := !!rhs)
```

#### Working with named lists

Another approach is to use named lists.  When a named list is passed to dplyr it converts it into a `:=` call.

```{r}
col <- "new_col"
val <- 3
expr <- setNames(list(rlang::expr(val)), col)
dplyr::mutate(mtcars, !!!expr)
```

Working with a named list is the preferred pattern when working with multiple programmatic columns:

```{r}
cols <- c("new_1", "new_2")
vals <- c(rlang::expr(carb * 2), rlang::expr(cyl / 2))
## vals is a vector of expressions to demonstrate how the concept can be applied to complex cases
expr <- setNames(vals, cols)
dplyr::mutate(mtcars, !!!expr)
```

::: {.callout-important}
`:=` is not a normal function; it is interpreted by dplyr to assign new column names dynamically.
:::

### Working with Columns

Working with columns requires building an expression for the righthand side of the `mutate()` function.

```{r}
dplyr::mutate(mtcars, new_col = wt / hp)
```

```json
{
  "mutate": {
    "new_col_name": "new_col",
    "operator": "/",
    "lhs": "wt",
    "rhs": "hp"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(mutate=list(new_col_name="new_col", "operator"="/", lhs="wt", rhs="hp")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r, filename="building an expression"}
build_mutate <- function(x) {
  rlang::call2(x$operator, rlang::sym(x$lhs), rlang::sym(x$rhs))
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_mutate(inp$mutate)
data <- dplyr::mutate(mtcars, !!inp$mutate$new_col_name := !!expr)
print(data)
```

```{r, filename="using a named list"}
build_mutate_list <- function(x) {
  setNames(
    list(rlang::call2(x$operator, rlang::sym(x$lhs), rlang::sym(x$rhs))),
    x$new_col_name
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_mutate_list(inp$mutate)
data <- dplyr::mutate(mtcars, !!!expr)
print(data)
```

### Generic Approach for Operators, Scalars and Columns {#sec-mutate-generic}

Based on what was discussed in @sec-filter-mapping-operators and @sec-combining-values-and-column-names, the mutate function can be extended to work with different operators and a mixture of scalar and column values.  For example, if we wish to calculate `new_col = mpg * cyl + 5`.

```{r}
dplyr::mutate(mtcars, new_col = mpg * cyl + 5)
```

From a conceptual point of view, it helps to work backwards from the `dplyr::mutate()` expression through the `rlang::call2()` expression and finally to the JSON format.  We'll then construct a generalized approach.

The `dplyr::mutate()` expression `mpg * cyl + 5` can be expressed as

```
"+"(
  "*"(
    mpg,
    cyl
  ),
  5
)
```

which, built using `rlang::call2()` would look like:

```{r}
#| eval: false
expr <- rlang::call2(
  "+", 
  rlang::call2(
    "*", 
    rlang::sym("mpg"), 
    rlang::sym("cyl")
  ), 
  5
)
```

To include this in a JSON expression, we can consider recursively breaking the entire expression into smaller and smaller expressions comprising of an operator, a left-hand side and a right-hand side (`lhs`:`operator`:`rhs`).  As an instruction set this could look like:

```json
{
  "new_col_name": "new_col",
  "operator": "+",
  "lhs": {
    "operator": "*",
    "lhs": "mpg",
    "rhs": "cyl"
  },
  "rhs": 5
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(new_col_name="new_col", operator="+", lhs=list(operator="*", lhs="mpg", rhs="cyl"), rhs=5), 'data/myfile.json', auto_unbox = TRUE)
```

and the engine would be:

```{r}
#| message: false
build_mutate <- function(x, col_names = c()) {

  if (
    is.character(x) && 
    length(x) == 1 && 
    x %in% col_names
  ) {
    return(rlang::sym(x))
  }

  if (
    is.atomic(x) && 
    length(x) == 1
  ) {
    return(x)
  }

  rlang::call2(x$operator,
    build_mutate(x$lhs, col_names),
    build_mutate(x$rhs, col_names)
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_mutate(inp, names(mtcars))
data <- dplyr::mutate(mtcars, !!inp$new_col_name := !!expr)
print(data)
```

The recursive function `build_mutate()` is similar in nature to the `build_filter()` function (@sec-filter-and-or).  `build_mutate()` takes an expression and keeps calling itself on the lefthand and righthand sides.  As with the `build_filter()` function, this is not foolproof but works for many general cases.  Ideally, the logic should be constructed so that anything that needs to be looked up at execution time is represented as a symbol, while anything else that should be taken as-is is represented as a literal.  To achieve this, the system needs to know when an input string represents a name to be evaluated versus when it represents a character value.  For example, it is impossible for this code to determine if "mpg" refers to the column named mpg or the literal string "mpg".  As in , one way to resolve this ambiguity is to encode the intent directly into the JSON format itself, for example:

```json
{
  "new_col_name": "new_col",
  "operator": "+",
  "lhs": {
    "operator": "*",
    "lhs": {
      "type": "column",
      "val": "mpg"
    },
    "rhs": {
      "type": "column",
      "val": "cyl"
    }
  },
  "rhs": {
    "type": "literal",
    "value": 5
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(new_col_name="new_col", operator="+", lhs=list(operator="*", lhs=list(type="column", val="mpg"), rhs=list(type="column", value="cyl")), rhs=list(type="literal", val=5)), 'data/myfile.json', auto_unbox = TRUE)
```

and the engine becomes

```{r}
#| message: false
build_mutate <- function(x) {

  if (!is.null(x$type)) {
    if (x$type == "column") {
      return(rlang::sym(x$val))
    }
    if (x$type == "literal") {
      return(x$val)
    }
  }

  rlang::call2(x$operator,
    build_mutate(x$lhs),
    build_mutate(x$rhs)
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_mutate(inp)
data <- dplyr::mutate(mtcars, !!inp$new_col_name := !!expr)
print(data)
```

Operators used in `mutate()` can be mapped in a similar manner to those for `filter()`.  For `mutate()` our operator mapping function could be:

```{r}
mutate_operators <- c(
  "+" = "+",
  "-" = "-",
  "*" = "*",
  "/" = "/",
  "log" = "log",
  "sqrt" = "sqrt"
)
```

This mapping allows the engine to convert operator strings in JSON into calls inside R, using `rlang::call2()`.  Our engine now becomes

```{r}
#| message: false
build_mutate <- function(x) {

  if (!is.null(x$type)) {
    if (x$type == "column") {
      return(rlang::sym(x$val))
    }
    if (x$type == "literal") {
      return(x$val)
    }
  }

  rlang::call2(mutate_operators[x$operator],
    build_mutate(x$lhs),
    build_mutate(x$rhs)
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

expr <- build_mutate(inp)
data <- dplyr::mutate(mtcars, !!inp$new_col_name := !!expr)
print(data)
```

## Summarizing Data

### `dplyr::summarize()`

While `filter()` and `mutate()` operate at the row level, processing each row indeividually, `dplyr::summarize()` (or `dplyr::summarise()`) is used to reduce the number of rows by computing aggregate values.  Each summary expression combines multiple rows into a single value, producing one output row per group (or single row if no grouping is applied).

### Grouping

#### Grouping using `group_by()`

The dplyr verb `group_by()` can be used to group rows by one or more variables.

```{r}
mtcars |>
  dplyr::group_by(cyl) |>
  dplyr::summarize(avg_mpg = mean(mpg))
```

Here `dplyr::group_by(cyl)` defines the groupoing metadata - data are grouped into matching values for **cyl**.  `dplyr::summarize()` evaluates the expression `avg_mpg = mean(mpg)` once per group and the result contains one row per value of **cyl**.  Grouping is persistent and the data remains grouped after summarization unless specifically dropped.

#### Grouping using `.by`

More recent versions of `dplyr` allow grouping to be specified within the `summarize()` verb directly by using the `.by` argument.

```{r}
mtcars |>
  dplyr::summarize(avg_mpg = mean(mpg), .by = cyl)
```

The differs from the example that uses `group_by()` in that grouping only applies to this call and the input data are not permanently grouped.

Both `.by` and `group_by()` support multiple columns.

```{r}
mtcars |>
  dplyr::summarize(avg_mpg = mean(mpg), .by = c(cyl, gear))
```

### Simple Summary Step

For a simple example, we will calculate the average mpg grouped by cyl and gear.

```{r}
dplyr::summarize(mtcars, avg_mpg = mean(mpg), .by = c(cyl, gear))
```

```json
{
  "summarize": {
    "by": ["cyl", "gear"],
    "metrics": {
      "name": "avg_mpg",
      "operator": "mean",
      "column": "mpg"
    }
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(summarize=list(by=list("cyl","gear"), metrics=list(name="avg_mean", operator="mean", column="mpg"))), 'data/myfile.json', auto_unbox = TRUE)
```

As we construct the engine it's clear that we will be dynamically assigning a new column name, in a manner similar to `mutate()` and therefore a named list should be used.

```{r}
build_summarize <- function(x) {
  setNames(
    list(rlang::call2(x$operator, rlang::sym(x$column))),
    x$name
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

## summarize expression
summarize_expr <- build_summarize(inp$summarize$metrics)

## list of grouping columns
by_expr <- rlang::syms(inp$summarize$by)

data <- dplyr::summarize(mtcars, !!summarize_expr, .by = c(!!!by_expr))
print(data)
```

In this implementation, the summarize expression is built using `build_summarize()`

```{r}
build_summarize(inp$summarize$metrics)
```

and the list of grouping columns is a list of syms

```{r}
rlang::syms(inp$summarize$by)
```

::: {.callout-tip}
When passing to the `summarize()` function, the `.by` argument expects a single. object (eg a vector or list).  Using `!!!by_expr` alone would splice two arguments (cyl and gear).  Passing `c(!!!by_expr)` collects them into a single object.
:::

### Multiple Summary Metrics

A single `summarize()` call often computes mutliple aggregates at once.  The output is one row per group with mutiple columns.

```{r}
dplyr::summarize(mtcars, avg_mpg = mean(mpg), max_hp = max(hp), n = dplyr::n(), .by = cyl)
```

```json
{
  "summarize": {
    "by": ["cyl"],
    "metrics": [
      { "name": "avg_mpg","operator": "mean", "column": "mpg"},
      { "name": "max_hp", "operator": "max", "column": "hp"},
      { "name": "n", "operator": "n" }
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(summarize=list(by=list("cyl"), metrics=list(list(name="avg_mean", operator="mean", column="mpg"), list(name="max_hp", operator="max", column="hp"), list(name="n", operator="n")))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
build_summarize <- function(x) {
  lapply(x, function(y) {
    if (y$operator == "n") {
      expr <- rlang::call2("n", .ns = "dplyr")
    } else {
      expr <- rlang::call2(y$operator, rlang::sym(y$column))
    }
    setNames(list(expr), y$name)
  }) |>
    unlist(recursive = TRUE)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

## summarize expression
summarize_expr <- build_summarize(inp$summarize$metrics)

## list of grouping columns
by_expr <- rlang::syms(inp$summarize$by)

data <- dplyr::summarize(mtcars, !!!summarize_expr, .by = c(!!!by_expr))
print(data)
```

Here, each summarize expression is built using `build_summarize()`

```{r}
build_summarize(inp$summarize$metrics)
```

and passed to `dplyr::summarize()` using `!!!`.  The list of grouping columns is, once again, a list of syms

```{r}
rlang::syms(inp$summarize$by)
```

## Multiple Columns in a Single Metric

A summary aggregate can depend on multiple columns.

```{r}
dplyr::summarize(mtcars, av_ratio = mean(mpg / hp), .by = cyl)
```

```json
{
  "summarize": {
    "by": ["cyl"],
    "metrics": [
      {
        "name": "avg_ratio",
        "operator": "mean",
        "expression": {
          "operator": "/",
          "lhs": "mpg",
          "rhs": "hp"
        }
      }
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(summarize=list(by=list("cyl"), metrics=list(list(name="avg_ratio", operator="mean", expression=list(operator="/", lhs="mpg", rhs="hp"))))), 'data/myfile.json', auto_unbox = TRUE)
```

The `build_expr()` function is the same as `build_mutate()` in @sec-mutate-generic

```{r}
build_expr <- function(x, col_names = c()) {

  if (
    is.character(x) && 
    length(x) == 1 && 
    x %in% col_names
  ) {
    return(rlang::sym(x))
  }

  if (
    is.atomic(x) && 
    length(x) == 1
  ) {
    return(x)
  }

  rlang::call2(x$operator,
    build_expr(x$lhs, col_names),
    build_expr(x$rhs, col_names)
  )
}

build_summarize <- function(x, col_names = c()) {
  lapply(x, function(y) {
    setNames(
      list(
        rlang::call2(y$operator, build_expr(y$expression, col_names))
      ),
      y$name
    )
  }) |>
    unlist(recursive = TRUE)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

## summarize expression
summarize_expr <- build_summarize(inp$summarize$metrics, names(mtcars))

## list of grouping columns
by_expr <- rlang::syms(inp$summarize$by)

data <- dplyr::summarize(mtcars, !!!summarize_expr, .by = c(!!!by_expr))
print(data)
```

### Using Parentheses in Summary Expressions

Operator precedence matters inside summary steps and therefore it is important to include a way to build up hierarchical expressions.

```{r}
dplyr::summarize(mtcars, score = mean((mpg + hp) / wt), .by = cyl)
```

From a conceptual approach, this can be considered as

```
mean(
  "/"(
    "+"(mpg, hp),
    wt
  )
)
```

and represented in JSON as

```json
{
  "summarize": {
    "by": ["cyl"],
    "metrics": [
      {
        "name": "score",
        "operator": "mean",
        "expression": {
          "operator": "/",
          "lhs": {
            "operator": "+",
            "lhs": "mpg",
            "rhs": "hp"
          },
          "rhs": "wt"
        }
      }
    ]
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(summarize=list(by=list("cyl"), metrics=list(list(name="score", operator="mean", expression=list(operator="/", lhs=list(operator="+", lhs="mpg", rhs="hp"), rhs="wt"))))), 'data/myfile.json', auto_unbox = TRUE)
```

The structure of the expression tree preserves the notion of parentheses; the order of operations is determined by how the expression is nested.

```{r}
build_expr <- function(x, col_names = c()) {

  if (
    is.character(x) && 
    length(x) == 1 && 
    x %in% col_names
  ) {
    return(rlang::sym(x))
  }

  if (
    is.atomic(x) && 
    length(x) == 1
  ) {
    return(x)
  }

  rlang::call2(x$operator,
    build_expr(x$lhs, col_names),
    build_expr(x$rhs, col_names)
  )
}

build_summarize <- function(x, col_names = c()) {

  lapply(x, function(y) {
    inner_expr <- build_expr(y$expression, col_names)
    summary_expr <- rlang::call2(y$operator, inner_expr)

    setNames(list(summary_expr), y$name)
  }) |>
    unlist(recursive = TRUE)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

## summarize expression
summarize_expr <- build_summarize(inp$summarize$metrics, names(mtcars))

## list of grouping columns
by_expr <- rlang::syms(inp$summarize$by)

data <- dplyr::summarize(mtcars, !!!summarize_expr, .by = c(!!!by_expr))
print(data)
```
