# Introduction

Once imported, one or more data wrangling steps are often required to manipulate data into its required form and format.  This section illustrates how data wrangling can be performed using an instruction set built on the {dplyr} and {tidyr} libraries.  Of course, other approaches, such as using base R functions and {data.table} may also be applied and the concepts presented here will hold true.

## Non-Standard Evaluation

In base R, most functions use standard evaluation, i.e. arguments are evaluated before the function is called.  For example when evaluating `mean(x)`, `x` is evaluated immediately and passed to the `mean()` function.

Many **tidyverse** verbs behave differently.  In functions such as `dplyr::filter()` and `dplyr::mutate()`, expressions are first captured and evaluated later, a process known as **non-standard evaluation (NSE)**.  For example in `dplyr::filter(mtcars, mpg > 30)`, the expression **mpg > 30** is not evaluated at the time the function is called.  Instead it is captured and evaluated later.  This delayed evaluation allows column names, such as **mpg**, to be written as though they were ordinary variables.

## Tidy Evaluation

Non-standard evaluation works well for interactive use but creates challenges when working programmatically.  Tidy evaluation addresses this by allowing expressions to be captured and manipulated as data.  One of the key building blocks of tidy evaluation is the *symbol*, which represents a name that is looked up when the expression is evaluated, rather than when it is created.  Symbols can be constructed programmatically using `rlang::sym()`.  For example, consider the three code blocks below:

```{r, filename="non-standard evaluation"}
## non-standard evaluation
## this code works
dplyr::filter(mtcars, mpg > 30)
```

```{r, filename="variables: fail"}
#| error: false
## this code fails
col <- "mpg"
value <- 30
dplyr::filter(mtcars, col > value)
```

```{r, filename="variables: success"}
## this code works
col <- "mpg"
expr <- rlang::sym(col)
value <- 30
dplyr::filter(mtcars, !!expr > value)
```

The first example works because it is the simplest case of non-standard evaluation and is well suited for interactive use.  
The second example attempts to run the same code using variables.  This fails because **col** in the expression **col > value** is treated as the literal variable name rather than being resolved to the column name **mpg**.  
In the third example, we construct an expression, `expr` from the *symbol* corresponding to **col**.  We then use the `!!` operator to inject that symbol into `dplyr::filter()`.  Since `rlang::sym(col)` produces the symbol **mpg**, the expression  
`dplyr::filter(mtcars, !!expr > value)`  
is equivalent to  
`dplyr::filter(mtcars, mpg > value)`.

While `!!` allows a single symbol or expression to be injected into a tidy-evaluated call, there are times where we wish to inject multiple expressions at once.  The `!!!` operator is used to splice a list of expressions into a function call.  This can be particularly useful when building a list of `dplyr::filter()` or `dplyr::arrange()` conditions.

## Data Masking

Many tidyverse verbs evaluate expressions within a **data mask**.  A data mask is an environment in which the columns of a data frame are made temporarily available as variables.  This allows expressions such as `mpg > 30` to be written without explicitly referencing the data frame that contains `mpg`.

For example, in the expression

```{r}
#| eval: false
dplyr::filter(mtcars, mpg > 30)
```

the symbol `mpg` is resolved by first looking it up in the data mask created from `mtcars`.  If a column with that name exists, it is used in the evaluation of the expression.

Data masking is closely related to non-standard evaluation but is conceptually different.  Non-standard evaluation describes *when* expressions are captured and evaluated, whereas data masking describes *where* names are resolved when the evaluation occurs.

## Argument Interfaces in the Tidyverse

Although many tidyverse verbs rely on tidy evaluation, they do not all accept arguments in the same way.  They do, however, expose different types of interfaces, each suited to particular classes of data manipulation tasks.  When working with the tidyverse programmatically, it is important to recognize that arguments are not interchangeable expressions, and that different interfaces require different forms of representation and processing.  The most commonly encountered interfaces can be grouped into three broad categories: *Tidy-Select Interfaces*, *Data-Masking Interfaces* and *Key-Based Interfaces*.

### Tidy-Select Interfaces

Functions such as `dplyr::select()`, `dplyr::rename()`, and parts of `tidyr::pivot_longer()` and `tidyr::pivot_wider()` use **tidy-select** semantics.  They do not accept expressions but operate on column specifications.  Tidy-select expressions support helpers such as `starts_with()`, `ends_with()`, `all_of()` and `any_of()`.  Although tidy-select expressions resemble data-masked expressions, they are evaluated using a separate selection mechanism.  Therefore, when building a programmatic interface, tidy-select arguments must be processed differently to data-masked ones.

::: {.callout-important}
While tidy-select was designed to support column selection, programmatic interfaces typically operate on explicit column names supplied as character vectors.  To support this, tidy-select provides helpers such as `all_of()` and `any_of()`.

These helpers convert character vectors into tidy-select specifications.  When building a selection from externally supplied column names, `all_of()` ensures that all column names exist, while `any_of()` silently ignores missing columns.

Unlike symbol injection (`!!sym()` or `!!!syms()`), tidy-select helpers do not rely on data-masking and not not construct executable expressions.  Instead they produce column selection specifications.

Tidy-select helpers provide a bridge between structured inputs, such as JSON, and tidy-select interfaces, without resorting to expression construction on non-standard evaluation.
:::

### Data-Masking Interfaces

Functions such as `dplyr::filter()`, `dplyr::mutate()`, `dplyr::summarise()`, and `dplyr::arrange()` accept **expressions** that are evaluated within a data-masked context.  These expressions may refer to column names, constants and functions, and are evaluated later using tidy evaluation.  Because these arguments represent executable expressions, they require special handling when constructed programmatically.  Typically, they are constructed using symbols (`sym()`, `syms()`) and spliced using `!!` and `!!!`.

### Key-Based Interfaces

Join verbs, such as `dplyr::left_join()`, `dplyr::inner_join()` or `dplyr::anti_join()` rely on key-based interfaces.  These functions operate on explicit column mappings rather than executable expressions.  When working with a JSON-type interface, these arguments require little or no tidy evaluation, making join operations comparatively simple to represent.

## Command Parsing {#sec-command-parsing}

The simplest approach is to specify a data manipulation step and execute it by evaluating it as an expression.  The base R functions `eval` and `parse` or the {rlang} function `parse_expr` can be used for this purpose.  In the example code below we define `filter_process`, a filter, as `mpg > 30` and evaluate it within a `dplyr::filter()` function.

```json
{
  "filter_process": "mpg > 30"
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(filter_process="mpg > 30"), 'data/myfile.json', auto_unbox = TRUE)
```

```{r filename="base R eval and parse"}
#| message: false
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::filter(mtcars, eval(parse(text = inp$filter_process)))
print(data)
```

```{r filename="rlang parse_expr"}
#| message: false
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::filter(mtcars, !!rlang::parse_expr(inp$filter_process))
print(data)
```

:::{.callout-important}
## Important
Such approaches are simplistic to build but open to abuse.  Parsing raw commands through an engine can often lead to undesired effects and should be avoided where possible.  An alternative approach is to break down the data wrangling steps.
:::

The following chapters will demonstrate how data wrangling steps can be broken down and reconstructed so that `eval(parse())` is not necessary.
