# Selection and Ordering

## Selecting Columns

### `dplyr::select()`

`dplyr::select()` picks or drops columns based on their names or positions.  It returns a new data frame containing only the chosen columns.  The number of rows is not affected.

### Retaining Columns

`dplyr::select()` takes one or more expressions separated by commas to specify columns.  For example, below the columns *mpg*, *cyl* and *wt* are retained, whilst all other columns are dropped.

```{r}
dplyr::select(mtcars, mpg, cyl, wt)
```

A select step is relatively simple to implement since `dplyr::select()` operates on names.  A simple example of a JSON input and engine for column selection is shown below.

```json
{
  "cols": ["mpg", "cyl", "wt"]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(cols=list("mpg", "cyl", "wt")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_select <- function(x) {
  rlang::syms(x)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, !!!build_select(inp$cols))
print(data)
```

Note the use of `!!!` and `rlang::syms()` as we are working with a list of arguments.  In the code above, `build_select(inp$cols)` equates to:

```{r}
build_select(inp$cols)
```

and the expression

```{r}
#| eval: false
dplyr::select(mtcars, !!!build_select(inp$cols))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::select(mtcars, mpg, cyl, wt)
```

### Dropping Columns

`dplyr::select()` can be used to drop columns.  For example, below the columns *mpg*, *cyl* and *wt* are dropped, whilst all other columns are retained.

```{r}
dplyr::select(mtcars, -mpg, -cyl, -wt)
```

To remove columns we can build an expressions using `rlang::call2()` as follows: `rlang::call2("-", rlang::sym(col_name))`.  This creates the expression `-colname` which can be inserted into `dplyr::select()`.  For multiple columns we simply create a list and use `!!!` as follows:

```json
{
  "drop_cols": ["mpg", "cyl", "wt"]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(drop_cols=list("mpg", "cyl", "wt")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_dropcols <- function(x) {
  lapply(x, function(y) {
    rlang::call2("-", rlang::sym(y))
  })
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::select(mtcars, !!!build_dropcols(inp$drop_cols))
print(data)
```

When constructing a function to drop columns, *build_dropcols()* returns a list of unevaluated expressions.  Since this is a list, `!!!` is used to splice each element in the `dplyr::select()` function.  In the code above, `build_dropcols(inp$cols)` equates to:

```{r}
build_dropcols(inp$cols)
```

and the expression

```{r}
#| eval: false
dplyr::select(mtcars, !!!build_dropcols(inp$cols))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::select(mtcars, -mpg, -cyl, -wt)
```

## Keeping Unique Rows

### `dplyr::distinct()`

`dplyr::distinct()` is used to retain unique rows. It can be used in two manners, returning only the unique values in the specified columns or returning all rows, keeping  just the first row when there are multiple matches.

### Retuning Only Selected Columns

`dplyr::distinct()` works in a similar manner to `dplyr::select()`, taking one or more expressions separated by commas to specify columns.  For example, below the unique values from the columns *gear*, *carb* are returned.

```{r}
dplyr::distinct(mtcars, gear, carb)
```

A simple example of a JSON input and engine for this approach is shown below.  It is identical to the *select* example.

```json
{
  "distinct": ["gear", "carb"]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(distinct=list("carb", "gear")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_distinct <- function(x) {
  rlang::syms(x)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::distinct(mtcars, !!!build_distinct(inp$distinct))
print(data)
```

### Retuning All Columns

`dplyr::distinct()` identifies rows with unique values.  If more than one row is satisfied,it will return the first match.  It can return either selected columns, as demonstrated above, or, if argument `.keep_all=TRUE`, will return all columns.

```{r}
dplyr::distinct(mtcars, gear, carb, .keep_all = TRUE)
```

Our engine can easily be extended to account for the use of `.keep_all` as follows.


```json
{
  "distinct": {
    "cols": ["gear", "carb"],
    "keep_all": true
  }
  
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(distinct=list(cols=list("carb", "gear"), keep_all=TRUE)), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_distinct <- function(x) {
  rlang::syms(x)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::distinct(mtcars, !!!build_distinct(inp$distinct$cols), .keep_all = isTRUE(inp$distinct$keep_all))
print(data)
```

The change in code occurs at `dplyr::distinct()`, and does not affect the way expressions are built.  The use of `isTRUE` accounts for missing values (`isTRUE(NULL)` equates to `FALSE`).

## Renaming Columns

### `dplyr::rename()`

`dplyr::rename()` changes the names of existing columns using `new_name = old_name` syntax.

### Renaming a Single Column

`dplyr::rename()` works on names, in a similar manner to `dplyr::select()`.  Renaming a single column is demonstrated below.

```{r}
dplyr::rename(mtcars, miles_per_gallon = mpg)
```

In JSON format this can be represented as:

```json
{
  "rename": {
    "old_name": "mpg",
    "new_name": "miles_per_gallon"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(rename=list(old_name="mpg", new_name="miles_per_gallon")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_rename <- function(x) {
  rlang::set_names(
    list(rlang::sym(x$old_name)), 
    x$new_name
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::rename(mtcars, !!!build_rename(inp$rename))
print(data)
```

`rlang::set_names()` is used to build a named list and pass it as an expression to `dplyr::rename()`.  The code follows the same approach as `select()` and `distinct()`.

### Renaming Multiple Columns

The same concept can be used when extending to multiple columns.

```{r}
dplyr::rename(mtcars, miles_per_gallon = mpg, cylinders = cyl, weight = wt)
```

```json
{
  "rename": {
    "mpg": "miles_per_gallon",
    "cyl": "cylinders",
    "wt": "weight"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(rename=list(mpg="miles_per_gallon", cyl="cylinders", wt="weight")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_rename <- function(x) {
  rlang::set_names(
    lapply(names(x), rlang::sym),
    unname(x)
  )
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::rename(mtcars, !!!build_rename(inp$rename))
print(data)
```

*build_rename(inp$rename)* returns:

```{r}
build_rename(inp$rename)
```

and 

```{r}
#| eval: false
dplyr::rename(mtcars, !!!build_rename(inp$rename))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::rename(mtcars, miles_per_gallon = mpg, cylinders = cyl, weight = wt)
```

## Ordering Rows

### `dplyr::arrange()`

`dplyr::arrange()` orders rows of a data frame by the values of selected columns.  `dplyr::arrange()` can sort rows using ascending or descending order and generally ignores grouping.

### Order on a Single Column

Ordering (ascending) on a single column is quite straightforward to implement.

```{r}
dplyr::arrange(mtcars, carb)
```

The JSON structure and engine look as follows:

```json
{
  "arrange": "carb"
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(arrange="carb"), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_arrange <- function(x) {
  rlang::sym(x)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::arrange(mtcars, !!build_arrange(inp$arrange))
print(data)
```


### Order on Multiple Columns

The approach can easily be extended to account for ordering (ascending) on multiple columns.

```{r}
dplyr::arrange(mtcars, carb, mpg)
```

The JSON structure and engine look as follows:

```json
{
  "arrange": ["carb", "mpg"]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(arrange=list("carb", "mpg")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_arrange <- function(x) {
  rlang::syms(x)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::arrange(mtcars, !!!build_arrange(inp$arrange))
print(data)
```

### Ascending and Descending Orders

Building a generic arrange function to account for multiple ascending and descending orders takes a little thought.  We cannot apply more than one `dplyr::arrange()` step, as later steps would override earlier ones leading to the data frame being re-sorted each time.  `dplyr::arrange()` is executed with no knowledge of previous sorting.  The entire `dplyr::arrange()` expression must be built into a single expression.  The example below loses any ordering from **carb** once it is ordered by **mpg**.

```{r}
dplyr::arrange(mtcars, carb) |>
  dplyr::arrange(mpg)
```

The `dplyr::arrange()` expression can be built up within an `lapply()` loop, accounting for ascending or descending order.  For example, the following sorts by gear (ascending), followed by cyl (descending) and mpg (ascending)

```{r}
dplyr::arrange(mtcars, gear, dplyr::desc(cyl), mpg)
```

A JSON format and engine could look like:

```json
{
  "arrange" : [
    {
      "column": "gear",
      "direction": "asc"
    },
    {
      "column": "cyl",
      "direction": "desc"
    },
    {
      "column": "mpg"
    }
  ]}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(arrange=list(list(column="gear", direction="asc"), list(column="cyl", direction="desc"), list(column="mpg"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
#| message: false
build_arrange <- function(x) {
  lapply(x, function(y) {
    if (isTRUE(y[['direction']] == "desc")) {
      rlang::call2("desc", rlang::sym(y[['column']]), .ns = "dplyr")
    } else { 
      rlang::sym(y[['column']])
    }
  })
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
data <- dplyr::arrange(mtcars, !!!build_arrange(inp$arrange))
print(data)
```

`build_arrange()` takes a list for ordering.  Each element is a list containing a column name and optional direction for sorting with the default being *ascending*.  When sorted as *ascending** `sym(column)` is returned and when sorted as *descending* `desc(sym(column))` is returned.  The use of `.ns` in `rlang::call2()` ensures that the namespace `dplyr` is prepended to `desc()`.  `build_arrange(inp$arrange)` returns:

```{r}
build_arrange(inp$arrange)
```

and 

```{r}
#| eval: false
dplyr::arrange(mtcars, !!!build_arrange(inp$arrange))
```

is equivalent to writing

```{r}
#| eval: false
dplyr::arrange(mtcars, gear, dplyr::desc(cyl), mpg)
```
