# Reshaping

Whereas verbs such as `filter()`, `mutate()` and `summarize()` operate on values in rows, reshaping verbs change the structure of the data itself.  Reshaping  typically affects the number of rows and/or columns by reorganizing how the data are represented.  Two reshaping verbs are covered here, `pivot_longer()` and `pivot_wider()`, both from the {tidyr} package.  Reshaping verbs take both tidy-select and key-based arguments, with arguments that refer specifically to column names as tidy-select and others as key-based.

## Pivoting Longer

### `pivot_longer()`

`tidyr::pivot_longer()` transforms data from wide to long format by collapsing multiple columns into key-value pairs.  It can take a number of arguments but the following are commonly used:

  - `cols` - list of columns to reshape
  - `names_to` - name of the new key column
  - `values_to` - name of the new value column

### Pivoting Variables Longer

```{r}
tidyr::pivot_longer(mtcars, cols = c(mpg, hp, wt), names_to = "metric", values_to = "value")
```

Here we pivot on the variables **mpg**, **hp** and **wt**, combining the values into a single column (**value**) with the **metric** column used for labeling.  In the resulting data frame the number of columns has been reduced by one (from 11 to 10) and the number of rows increased by 64 (from 32 to 96).

We can introduce a JSON format for pivot_longer:

```json
{
  "pivot_longer": {
    "cols": ["mpg", "hp", "wt"],
    "names_to": "metric",
    "values_to": "value"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(pivot_longer=list(cols=list("mpg", "cyl", "wt"), names_to="metric", values_to="value")), 'data/myfile.json', auto_unbox = TRUE)
```

Since `pivot_longer()` uses just tidy-select and key-based arguments, no expression building is required and an engine to process `pivot_longer()` is relatively simple.

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

pivot_cols <- unlist(inp$pivot_longer$cols, use.names = FALSE)
data <- tidyr::pivot_longer(
    mtcars, 
    cols = tidyselect::all_of(pivot_cols), 
    names_to = inp$pivot_longer$names_to, 
    values_to = inp$pivot_longer$values_to
)
print(data)
```

See @sec-wrangling-anyof for more details on the use of tidy-select helpers.

## Pivoting Wider

### `pivot_wider()`

`tidyr::pivot_longer()` transforms data from long to wide format by spreading key-value pairs into separate columns.  It is the inverse transformation of `pivot_longer()`.  `pivot_wider()` can take a number of arguments but the following are commonly used:

  - `id_cols` - a set of columns that uniquely identify each observation.  Defaults to all columns (except those specified through `names_from` and `values_from`) if omitted.
  - `names_from` - column to use for the output column
  - `values_from` - column to use for the cell values

### Pivoting Variables Wider

```{r}
tidyr::pivot_wider(mtcars, id_cols = c(mpg, hp, wt), names_from = gear, values_from = carb)
```

Here, the columns **mpg**, **hp** and **wt** are retained, and they uniquely identify each row.  The **gear** column is transformed wide so that 3 new columns are created, one for each value of **gear**.  These new columns are populated with the values found in **carb**.

In JSON format:

```json
{
  "pivot_wider": {
    "cols": ["mpg", "hp", "wt"],
    "names_from": "gear",
    "values_from": "carb"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(pivot_wider=list(cols=list("mpg", "cyl", "wt"), names_to="gear", values_to="carb")), 'data/myfile.json', auto_unbox = TRUE)
```

Whereas `pivot_longer()` treats `names_to` and `values_to` as literal strings naming new columns, `pivot_wider()` treats `names_from` and `values_from` as tidy-select column identifiers.  As a result `pivot_wider()` requires tidy-select helpers.

```{r}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

pivot_cols <- unlist(inp$pivot_wider$cols, use.names = FALSE)
data <- tidyr::pivot_wider(
    mtcars, 
    id_cols = tidyselect::all_of(pivot_cols), 
    names_from = tidyselect::all_of(inp$pivot_wider$names_to), 
    values_from = tidyselect::all_of(inp$pivot_wider$values_to)
)
print(data)
```

### Additional Arguments

`pivot_wider()` can accept several additional optional arguments.  They are treated as key-based arguments and the engine can be extended to accommodate these.  Some examples are shown below.

#### `names_prefix`

`names_prefix` is used to add a string to the start of each new column name.

```{r}
tidyr::pivot_wider(mtcars, id_cols = c(mpg, hp, wt), names_from = gear, values_from = carb, names_prefix = "gear_")
```

In JSON format:

```json
{
  "pivot_wider": {
    "cols": ["mpg", "hp", "wt"],
    "names_from": "gear",
    "values_from": "carb",
    "names_prefix": "gear_"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(pivot_wider=list(cols=list("mpg", "cyl", "wt"), names_to="gear", values_to="carb", names_prefix="gear_")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

pivot_cols <- unlist(inp$pivot_wider$cols, use.names = FALSE)
data <- tidyr::pivot_wider(
    mtcars, 
    id_cols = tidyselect::all_of(pivot_cols), 
    names_from = tidyselect::all_of(inp$pivot_wider$names_to), 
    values_from = tidyselect::all_of(inp$pivot_wider$values_to),
    names_prefix = inp$pivot_wider$names_prefix
)
print(data)
```

#### `values_fill`

`values_fill` determines how to address missing values.  It can be a scalar or a named list if applying different values to each column.

```{r}
tidyr::pivot_wider(mtcars, id_cols = c(mpg, hp, wt), names_from = gear, values_from = carb, values_fill = 0)
```

In JSON format:

```json
{
  "pivot_wider": {
    "cols": ["mpg", "hp", "wt"],
    "names_from": "gear",
    "values_from": "carb",
    "values_fill": 0
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(pivot_wider=list(cols=list("mpg", "cyl", "wt"), names_to="gear", values_to="carb", values_fill=0)), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

pivot_cols <- unlist(inp$pivot_wider$cols, use.names = FALSE)
data <- tidyr::pivot_wider(
    mtcars, 
    id_cols = tidyselect::all_of(pivot_cols), 
    names_from = tidyselect::all_of(inp$pivot_wider$names_to), 
    values_from = tidyselect::all_of(inp$pivot_wider$values_to),
    values_fill = inp$pivot_wider$values_fill
)
print(data)
```

If `values_fill` is missing we can apply a detault value using the rlang `%||%` operator.

```json
{
  "pivot_wider": {
    "cols": ["mpg", "hp", "wt"],
    "names_from": "gear",
    "values_from": "carb"
  }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(pivot_wider=list(cols=list("mpg", "cyl", "wt"), names_to="gear", values_to="carb")), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
library(rlang)
default_missing <- -99

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)

pivot_cols <- unlist(inp$pivot_wider$cols, use.names = FALSE)
data <- tidyr::pivot_wider(
    mtcars, 
    id_cols = tidyselect::all_of(pivot_cols), 
    names_from = tidyselect::all_of(inp$pivot_wider$names_to), 
    values_from = tidyselect::all_of(inp$pivot_wider$values_to),
    values_fill = inp$pivot_wider$values_fill %||% default_missing
)
print(data)
```

::: {.callout-tip}
`%||%` is an infix function which works as follows:  
`x %||% y` If `x` is NULL then return `y` otherwise return `x`
:::
