# Expression Handling {#sec-expression-handling}

## Introduction

Expression handling is fundamental to working with data in R, particularly when using libraries such as dplyr and ggplot2.  This chapter covers how R captures, manipulates, and evaluates expressions.

Most R functions use *standard evaluation*, where arguments are evaluated immediately before the function executes.  Many *tidyverse* functions, however, use *non-standard evaluation* (NSE), which captures expressions unevaluated for downstream processing.

## Non-Standard Evaluation

### Standard Evaluation

In standard evaluation, arguments are evaluated before the function is called.  In this example:

```r
x <- c(1, 2, 3, 4, 5)
mean(x)
```

`x` is evaluated immediately and the result is passed to `mean()`.

### Non-Standard Evaluation in Data Manipulation

Many **tidyverse** verbs behave differently.  In functions such as `dplyr::filter()` and `dplyr::mutate()`, expressions are first captured and evaluated later, a process known as **non-standard evaluation (NSE)**.  For example in

```r
dplyr::filter(mtcars, mpg > 30)
```

the expression **mpg > 30** is not evaluated at the time the function is called.  Instead it is captured and evaluated in a special context where column names from `mtcars` are available as variables.  This allows `mpg` to be written without having to reference `mtcars` as you would in base R, `mtcars$mpg`.

### When to use Non-Standard Evaluation?

NSE provides a more intuitive syntax for interactive use.  Typing `filter(mtcars, mpg > 30)` is cleaner than `filter(mtcars, mtcars$mpg > 30)`.  NSE creates challenges, however, when working *programmatically*.  For example when building conditions dynamically or passing column names as strings.

## Tidy Evaluation

Tidy evaluation extends non-standard evaluation to make it composable and programmable.  It provides a consistent framework for capturing expressions and manipulating them as data.

### Symbols

A *symbol* is one of key building blocks of tidy evaluation.  It represents a name that is looked up when evaluated, rather than when it is created.  Symbols can be constructed programmatically using `rlang::sym()`.  For example:

```{r}
# Create a symbol from a string
col_name <- "mpg"
sym_col <- rlang::sym(col_name)
print(sym_col)
```

### Quasiquotation: The `!!` Operator

Once you have a symbol or expression, you can inject it into a function call using the `!!` (bang-bang) operator:

```{r}
library(rlang)

col_name <- "mpg"
threshold <- 30

# This will fail: col_name is treated as a literal variable name
dplyr::filter(mtcars, col_name > threshold)

# This works: the symbol is injected using !!
expr <- sym(col_name)
dplyr::filter(mtcars, !!expr > threshold)
```

The `!!` operator tells the function: "This is not a literal name; evaluate this expression and inject the result".

### Quasiquotation: The `!!!` Operator

To inject multiple expressions at once, the `!!!` (bang-bang-bang) operator can be used.  This splices a list of expressions into the function call.  

#### `expr()` vs `sym()`

Before diving into `!!!`, it is helpful to understand the difference between `expr()` and `sym()`.

-  **`sym(name)`** creates a *symbol*, which is a reference to a single variable or column name.  It can be used with `!!` to inject the name into a function call.  Operators, however, cannot be used in symbols.  `sym("mpg")` works but `sym(mpg) > 30` fails.
-  **`expr()`** captures an entire *expression* which can include operators, function calls and complex logic.  It is used when you need to capture more than just a name, for example `mpg > 30`.

Here is an example of using `!!!` to filter against a list of conditions.  The conditions are built using `expr()`.

```{r}
# Build a list of filter conditions
filter_conditions <- list(
    expr(mpg > 30),
    expr(gear == 5)
)

# Splice all the conditions into the filter function
dplyr::filter(mtcars, !!!filter_conditions)
```

### Combining Quasiquotation with Functions

To use quasiquotations inside functions you can use the `enquo()` function to capture an argument as a quosure (a quoted expression with its environment).  For example:

```{r}
summarize_by <- function(data, group_col, summary_col) {
    group_quo <- enquo(group_col)
    summary_quo <- enquo(summary_col)

    data |> 
        dplyr::group_by(!!group_quo) |>
        dplyr::summarize(mean = mean(!!summary_quo, na.rm = TRUE))
}

summarize_by(mtcars, cyl, mpg)
```

## Data Masking

Many tidyverse verbs evaluate expressions within a **data mask**.  A data mask is an environment in which the columns of a data frame are made temporarily available as variables.

### How Data Masking Works

Consider the expression

```r
dplyr::filter(mtcars, mpg > 30)
```

Inside `filter()`, the data mask makes `mpg` available as if it were a variable, even though it is actually a column.  This is possible because:

1. `mpg > 30` is capture and not evaluated immediately
2. `dplyr` creates an environment where `mpg` refers to the `mpg` column
3. The expression is evaluated in this masked environment
4. The result (a logical vector) is used to filter the rows of `mtcars`

Data masking only applies to certain contexts within tidyverse functions.  For example

```r
# Data masking applies here
# mpg is looked up in the data mask
dplyr::mutate(mtcars, new_col = mpg * 2)

# Data masking does not apply here
# We are assigning a string so nothing is evaluated
dplyr::mutate(mtcars, new_col = "miles_per_gallon")
```

::: {.callout-note}
Data masking is closely related to non-standard evaluation but is conceptually different.  Non-standard evaluation describes *when* expressions are captured and evaluated, whereas data masking describes *where* names are resolved when the evaluation occurs.
:::

### Pronouns: `.data` and `.env`

When using data masking sometimes you need to determine if a name should be looked up in the data or in the environment.  The `.data` and `.env` pronouns provide explicit control.

```{r}
#| eval: false
gear <- 6

# Ambiguous
# is threshold a column or an object?
dplyr::filter(mtcars, cyl > gear)
```

```{r}
gear <- 6

# Explicit
# gear is in the environment
dplyr::filter(mtcars, cyl > .env$gear)
```

```{r}
gear <- 6

# Explicit
# gear is in the data
dplyr::filter(mtcars, cyl > .data$gear)
```

## Practical Examples

Once you have got the hand of tidy evaluation and data masking, you can build complex, reusable functions for data manipulation.

### Dynamic Column Selection

Functions that will work with any column

```{r}
get_summary <- function(data, col_name) {
    col <- sym(col_name)

    data |> 
        dplyr::summarize(
            mean = mean(!!col, na.rm = TRUE),
            df = sd(!!col, na.rm = TRUE),
            n = dplyr::n()
        )
}

get_summary(mtcars, "mpg")
```

### Building Complex Filters Programmatically

Construct filter conditions from external specifications, for example user input or configuration file.

```{r}
apply_filters <- function(data, filter_specs) {
    conditions <- lapply(filter_specs, function(spec) {
        col <- sym(spec$col)
        op <- spec$op
        val <- spec$val

        switch(op,
            ">" = expr(!!col > !!val),
            "==" = expr(!!col == !!val),
            "<" = expr(!!col < !!val)
        )
    })
    dplyr::filter(data, !!!conditions)
}

filter_conditions <- list(
    list(col = "mpg", op = ">", val = 30),
    list(col = "gear", op = "==", val = 5)
)
apply_filters(mtcars, filter_conditions)
```

### Creating Reusable Transformation Functions

Build functions that transform data in flexible ways.

```{r}
normalize_column <- function(data, col_name) {
    col <- sym(col_name)

    data |> dplyr::mutate(!!col := (!!col - mean(!!col, na.rm = TRUE)) / sd(!!col, na.rm = TRUE))
}

normalize_column(mtcars, "mpg")
```

## When to Build Expressions vs Parse Strings

Expression handling can be implemented in two different ways, and the choice matters for safety, testability and maintainability.  Programmatic construction is a bottom-up approach that builds an expression from its parts.  String parsing is a top-down approach that takes an expression and safely evaluates it.

### Programmatic Construction

-  Use when inputs are *structured* (eg JSON or YAML with a dictionary of arguments) or when the system generates expressions.
-  Useful tools are `rlang::sym()`, `rlang::syms()`, `rlang::call2()`, `rlang::expr()`.  Splice with `!!` and `!!!`.
-  Advantages: it is explicit, testable, predictable and ultimately safer.  It would be the **default choice** when you want to control the input format.

### String Parsing

-  Use when users or external systems supply free-form expressions.
-  The general pattern would be to parse and validates the expression before evaluating.
-  Trade-offs are that it is flexible but requires strict validation, careful scoping and clear error handling.

### When to use Each

-  Structured input => **Programmatic construction**.
-  Free-text user input => **String parsing**.

The next section (@sec-expression-parser) provides a concrete approach to expression parsing.

## Building a Safe Expression Parser {#sec-expression-parser}

While quasiquotation tools are powerful for everyday programming, they may not be suitable when evaluating expressions that come from external sources.  In these cases a safer approach is often to parse, validate and safely execute the expressions.

1. **Parse** the expression into an Abstract Syntax Tree (AST) without executing.
2. **Validate** the AST to ensure only allowed operators are present.
3. **Evaluate** through data masking (tidyverse verbs) or in a restricted environment (for scalar expressions).

```{mermaid}
flowchart TD
    A(Parse) --> B(Validate)
    B --> C(Evaluate)
```

### Understanding the Abstract Syntax Tree (AST)

An abstract syntax tree is a tree-structured representation of an expression.  

The `ast()` function from the {lobstr} package can be used to convert an expression into an AST.  For example:

```{r}
lobstr::ast(1 + 2)
lobstr::ast(mean(set1, set2, set3))
lobstr::ast(sd(c(4, 5, 6)))
lobstr::ast(12 + (9 - x * (4 - y) ^ 2))
```

More information on expressions and the abstract syntax trees can be found in *Advanced R*.[^1]

[^1]: Wickham, H. (2019). Advanced R (2nd ed.). Chapman & Hall/Routledge [https://adv-r.hadley.nz/expressions.html#ast](https://adv-r.hadley.nz/expressions.html#ast)

### A Safe Expression Parser for Data-Masking

The three steps for safely evaluating a data-masked expression are:

1. **Parse** the expression into an Abstract Syntax Tree (AST) without executing.
2. **Validate** the AST to ensure only allowed operators are present.
3. **Evaluate** through data masking.

#### Step 1. Parsing Expressions {#sec-validate-dm-parse}

`rlang::parse_expr()` parses an expression without executing it.

::: {.callout-tip}
`rlang::parse_expr()` parses an expression without running it.  Parsing an expression produces an AST (abstract expression tree).  For example, running `rlang::parse_expr("1 + 2")` produces the AST

```{r}
#| echo: false
lobstr::ast(1 + 2)
```
:::

```{r}
#| error: true
rlang::parse_expr("1 + 2")
rlang::parse_expr("age > 30")
rlang::parse_expr("42")
# This will error as `Hello World` is not an expression
rlang::parse_expr("Hello World")
rlang::parse_expr("\"Hello World\"")
```

#### Step 2. Validating the AST {#sec-validate-dm-ast}

Validation walks the AST recursively and rejects any disallowed operators.  Nothing is evaluated at this stage; the function just verifies that only a small, approved subset of R operators are used.

Let's start by setting a set of allowed operators and functions.  When we validate an expression we check each function (in R operators such as `+` and `-` are functions).

```{r}
allowed_operators <- c("+", "-", "*", "/", "^", 
                       ">", ">=", "<", "<=", "==", "!=", 
                       "&", "|", "&&", "||", "!",
                       "c", "mean", "sd")
```

To validate the AST we'll build a function that returns `TRUE` if the AST is valid and stops if it is not.  A more sophisticated approach would return an error and allow the logic in the engine to determine what to do.

The function below validates an expression for data masking.

```{r}
validate_expr <- function(expr) {
  
  # Atomic values (numbers, strings, logicals) are always safe
  if (rlang::is_atomic(expr)) {
    return(TRUE)
  }

  # Symbols (variables and column names) are safe
  if (rlang::is_symbol(expr)) {
    return(TRUE)
  }

  # Check function calls
  if (rlang::is_call(expr)) {
    fn <- rlang::call_name(expr)

    # Parentheses are grouping operators, not function calls
    if (fn == "(") {
      validate_expr(expr[[2]])
      return(TRUE)
    }

    # Reject disallowed operators
    if (!fn %in% allowed_operators) {
      stop("Disqualified operator: ", fn, call. = FALSE)
    }

    # Recursively validate arguments
    for (i in 2:length(expr)) {
      validate_expr(expr[[i]])
    }

    return(TRUE)
  }

  # Anything else is rejected
  stop("Unsupported expression type", call. = FALSE)
}
```

The `validate_expr()` function walks through the parsed expression (AST) and ensures that **every** part is allowed before anything is evaluated.  Only simple arithmetic, logical expressions and a small subset of functions, as defined in `allowed_operators`, but nothing else is allowed.  If the AST is composed of safe operators and values then it returns `TRUE`, if not then it **stops**.  The function works by performing a recursively traversing the AST with the following logic:

-  If the node is an atomic value (eg numeric, logical or string) then return `TRUE`.  All atomics are safe as they cannot execute code.
-  If the node is a variable or column name then return `TRUE`.
-  If the node is a function call (eg `1 + 2`, `x > 5`) then
    -  Extract the operator/function name (`fn`).
    -  Check for parentheses and validate the inner expression.  See @nte-expressions-parentheses.
    -  check that `fn` is allowed by confirming it is in `allowed_operators` list.  The ensures that **only** the subset of operators or functions can be used.  If the operator is not in the `allowed_operators` list then stop.
    -  recursively validate each argument in the node.
-  Anything else is rejected.

The following examples illustrate that, although `max(c(1,2,3))` is a valid R expression, it is rejected by `validate_expr()` as `max()` is not in `allowed_calls`.

```{r}
#| error: true
# valid expression
validate_expr(rlang::parse_expr('1 + 2'))
# valid expression
validate_expr(rlang::parse_expr('mean(c(1,2,3))'))
# invalid expression - `max` is not allowed
validate_expr(rlang::parse_expr('max(c(1,2,3))'))
# invalid expression - `[`] is not allowed
validate_expr(rlang::parse_expr('x[1]'))
# invalid expression - `library` is not allowed
validate_expr(rlang::parse_expr('library(dplyr)'))
```

::: {#nte-expressions-parentheses .callout-note}
##### Parentheses in AST
Parentheses are ignored.  They are not function calls, but are grouping.  The AST for the expression `x > 1` is

```{r}
lobstr::ast(x > 1)
```

whereas the AST for the expression `(x > 1)` is

```{r}
lobstr::ast((x > 1))
```

This part of `validate_expr()` ignores the parentheses and validates the inner part (in this case `>`(x, 1)) of the expression ony:

```{r}
#| eval: false
if (fn == "(") {
  validate_expr(expr[[2]])
  return(TRUE)
}
```
:::

#### Step 3. Evaluation Through Data Masking

The validated expression can now be safely injected into tidyverse verbs, which then evaluate it using data masking.  For example using `dplyr::filter()`

```{r}
expr <- rlang::parse_expr("mpg > 30")
validate_expr(expr)
dplyr::filter(mtcars, !!expr)
```

or `dplyr::mutate()`

```{r}
expr <- rlang::parse_expr("mpg ^ 2")
new_col_name <- "new_col"
validate_expr(expr)
dplyr::mutate(mtcars, !!new_col_name := !!expr)
```

This approach will fail as `max()` is not included in the list of allowed functions and operators.  Whether functions such as `max()` are permitted is entirely determined by the validation rules.

```{r}
#| error: true
expr <- rlang::parse_expr("max(mpg)")
new_col_name <- "new_col"
validate_expr(expr)
# dplyr::mutate(mtcars, !!new_col_name := !!expr)
```

The examples above inject the validated expression into a data-masked tidyverse verb.  The same expression can be used in base R if evaluated using `rlang::eval_tidy()`.

Injecting to a tidyverse verb that takes a data-mask as an argument:
```{r}
expr <- rlang::parse_expr("mpg > 30")
validate_expr(expr)
dplyr::filter(mtcars, !!expr)
```

Filtering with base R on the same expression:
```{r}
expr <- rlang::parse_expr("mpg > 30")
validate_expr(expr)
expr_eval_tidy <- rlang::eval_tidy(expr, data = mtcars)
mtcars[expr_eval_tidy, ]
```

`rlang::eval_tidy()` returns a vector of logicals that is simply passed into mtcars to filter rows passing the condition.

```{r}
print(expr_eval_tidy)
```

Adding a new column can be achieved as follows:
```{r}
d <- mtcars
expr <- rlang::parse_expr("mpg ^ 2")
new_col_name <- "new_col"
validate_expr(expr)
expr_eval_tidy <- rlang::eval_tidy(expr, data = d)
d[[new_col_name]] <- expr_eval_tidy
print(d)
```

where `expr_eval_tidy` evaluates to

```{r}
print(expr_eval_tidy)
```

### A Safe Expression Parser for Scalar Expressions

The three steps for safely evaluating a scalar expression are:

1. **Parse** the expression into an Abstract Syntax Tree (AST) without executing.
2. **Validate** the AST to ensure only allowed operators are present.
3. **Evaluate** in a restricted environment.

#### Step 1. Parsing Expressions

As explored in @sec-validate-dm-parse, `rlang::parse_expr()` parses an expression without executing it.  Refer to @sec-validate-dm-parse for more details. 

#### Step 2. Validating the AST

The `validate_expr()` function from @sec-validate-dm-ast works well for situations when our expression is being validated for inclusion into a data-mask.  There are, however, situations when we wish to evaluate expressions outside of data-masking.  For example using an expression in `dplyr::filter()` would require data-masking, whereas evaluating `mean(c(1, 2, 3))` would not.  We need to make one change to `validate_expr()` so that it works with scalar expressions: logic for **symbols** should be removed, since scalar evaluation is not performed within in a data mask.[^2]

[^2]: In contrast, expressions evaluated inside tidyverse verbs such as `filter()` or `mutate()` require symbols, as column names are resolved through a data mask.

As discussed in @sec-validate-dm-ast, validation walks the AST recursively and rejects any disallowed operators.  Nothing is evaluated at this stage; the function just verifies that only a small, approved subset of R operators are used.

The same set of allowed operators and functions are used:

```{r}
allowed_operators <- c("+", "-", "*", "/", "^", 
                       ">", ">=", "<", "<=", "==", "!=", 
                       "&", "|", "&&", "||", "!",
                       "c", "mean", "sd")
```

To validate the AST we'll build a new function, `validate_var_expr()`, that returns `TRUE` if the AST is valid and stops if it is not.  

```{r}
validate_var_expr <- function(expr) {
  
  # Atomic values (numbers, strings, logicals) are always safe
  if (rlang::is_atomic(expr)) {
    return(TRUE)
  }

  # Check function calls
  if (rlang::is_call(expr)) {
    fn <- rlang::call_name(expr)

    # Parentheses are grouping operators, not function calls
    if (fn == "(") {
      validate_var_expr(expr[[2]])
      return(TRUE)
    }

    # Reject disallowed operators
    if (!fn %in% allowed_operators) {
      stop("Disqualified operator: ", fn, call. = FALSE)
    }

    # Recursively validate arguments
    for (i in 2:length(expr)) {
      validate_var_expr(expr[[i]])
    }

    return(TRUE)
  }

  # Anything else is rejected
  stop("Unsupported expression type", call. = FALSE)
}
```

The `validate_var_expr()` function walks through the parsed expression (AST) and ensures that **every** part is allowed before anything is evaluated.  Only simple arithmetic, logical expressions and a small subset of functions, as defined in `allowed_operators`, but nothing else is allowed.  If the AST is composed of safe operators and values then it returns `TRUE`, if not then it **stops**. 

#### Step 3. Safely Evaluating the Expression

Now our expression contains only operators from the `allowed_operators` list we evaluate it.  A simple helper function combines validation and evaluation as follows:

```{r}
safe_eval <- function(expr) {
  validate_var_expr(expr)
  eval(expr)
}
```

```{r}
safe_eval(rlang::parse_expr('1 + 2'))
```

By default, `eval()` evaluates within the environment in which it was called.  Even if our AST contains just operators listed in `allowed_calls`, R still has to look up these symbols when evaluating an expression.  This means that symbol lookup is affected by the current environment, and even core operators such as `+` can be rebound:

```{r}
## define an expression for evaluation
my_expr <- rlang::parse_expr("9 + 5")

eval(my_expr)

## store default behavior of `+` and reassign `+` to act as `-`
add_operator_default <- `+`
`+` <- `-`

eval(my_expr)

## reset `+` back to original behavior
`+` <- add_operator_default
```

Wow!  We've just redefined how basic mathematics works in R.

To prevent user expressions from being affected by rebinding in the calling environment, we can evaluate them in a dedicated, safe environment:

```{r}
## build a safe environment for evaluation
safe_env <- new.env(parent = emptyenv())

rlang::env_bind(
  safe_env,
  `+`  = `+`,
  `-`  = `-`,
  `*`  = `*`,
  `/`  = `/`
)

## store default `+` behavior
add_operator_default <- `+`

## define an expression for evaluation
my_expr <- rlang::parse_expr("9 + 5")

eval(my_expr, envir = safe_env)

## reassign `+` to act as `-`
`+` <- `-`

eval(my_expr, envir = safe_env)

## reset `+` back to original behavior
`+` <- add_operator_default
```

Here we have create a new environment, `safe_env`, which is built on the empty environment.  The empty environment contains no operators or functions, not even primitive operators such as `+` or `*`.  `rlang::env_bind()` allows us to create bindings, essentially defining functions and operators for our environment.  We have used `rlang::env_bind()` to add basic mathematical functions to `safe_env`.  Now when we rebind `+` it does not have an effect on evaluating our expression since it is evaluated in `safe_env`.

The beauty of this approach is that we have full control over what is permitted in our environment.

When we evaluate using `eval()` we supply `safe_env` which contains only mappings for our functions and operators.  We can confirm that functions not listed in our environment are blocked.

The following code will fail since both `c()` and `mean()` are missing from our safe environment.

```{r}
#| error: true
## define an expression for evaluation
my_expr <- rlang::parse_expr("mean(c(1, 2, 3))")

## evaluate in global environment
eval(my_expr)

## evaluate in safe environment
eval(my_expr, envir = safe_env)
```

After updating our environment `mean(c(1, 2, 3))` can be evaluated.

```{r}
safe_env <- new.env(parent = emptyenv())
rlang::env_bind(
  safe_env,
  `+`  = `+`,
  `-`  = `-`,
  `*`  = `*`,
  `/`  = `/`,
  `^`  = `^`,
  `>`  = `>`,
  `<`  = `<`,
  `>=` = `>=`,
  `<=` = `<=`,
  `==` = `==`,
  `!=` = `!=`,
  `&`  = `&`,
  `|`  = `|`,
  `&&` = `&&`,
  `||` = `||`,
  `!`  = `!`,
  `(` = `(`,
  c = c,
  mean = mean,
  sd = sd
)

safe_eval <- function(expr) {
  validate_var_expr(expr)
  eval(expr, envir = safe_env)
}
```

```{r}
#| error: true
## define an expression for evaluation
my_expr <- rlang::parse_expr("mean(c(1, 2, 3))")

## evaluate in global environment
eval(my_expr)

## validate and evaluate in safe environment
safe_eval(my_expr)
```

### A Single Function For Safe Expression Handling {#sec-validate-function}

We can build a wrapper function for parsing, validating and evaluating for both data-mask and scalar expressions.  

```{r}
eval_safe_expr <- function(
  expr_input,
  mode = c("inject", "evaluate"),
  eval_context = c("data_mask", "scalar"),
  data = NULL,
  validator = NULL
) {

  mode <- match.arg(mode)
  eval_context <- match.arg(eval_context)

  # Parse if a string is supplied
  expr <- if (is.character(expr_input) && length(expr_input) == 1) {
    rlang::parse_expr(expr_input)
  } else {
    expr_input
  }

  # If validator is not specified then choose it based on eval_context
  if (is.null(validator)) {
    validator <- if (eval_context == "data_mask") {
      validate_expr
    } else {
      validate_var_expr
    }
  }

  # Validate
  # Will stop on failure
  validator(expr)

  # mode: inject
  # Return validated expression
  if (mode == "inject") {
    return(expr)
  }

  # mode: evaluate
  # Evaluate the expression
  if (eval_context == "data_mask") {
    # Evaluate data mask using eval_tidy
    if (is.null(data)) stop("Data Required")
    return(rlang::eval_tidy(expr, data = data))
  } else {
    # Safely evaluate scalar expression
    eval(expr, envir = safe_env)
  }
}
```

Approach for tidyverse verbs:
```{r}
expr <- eval_safe_expr("mpg > 30", mode = "inject", eval_context = "data_mask")
dplyr::filter(mtcars, !!expr)
```

Evaluate outside a verb:
```{r}
logical_vec <- eval_safe_expr("mpg > 30", mode = "evaluate", eval_context = "data_mask", data = mtcars)
mtcars[logical_vec, ]
```

Scalar evaluation within a restricted environment:
```{r}
expr <- "mean(c(1, 2, 3)) + 10"
eval_safe_expr(expr, mode = "evaluate", eval_context = "scalar")
```

Scalar evaluation with an unrecognized function:
```{r}
#| error: true
expr <- "sum(c(1, 2, 3)) + 10"
eval_safe_expr(expr, mode = "evaluate", eval_context = "scalar")
```
