# A Generic Approach to Visualizations {#sec-visualizations-generic}

In this chapter we'll discuss a generic approach to static chart building with ggplot2.  Starting with a specific approach, demonstrating with a scatter chart, we'll move on to how this can be built as a generic approach, applicable to any ggplot2 chart with geoms.

## Building Specific Charts

We'll start with building a basic scatter chart from an instruction set.

The simplest form is to just import a column identifiers for `x` and `y`, and call `geom_point()`.  Assuming our parameters are available in variables `x` and `y` and `data` contains a data frame to be plotted, a scatter chart can be drawn as follows:

```{r}
data <- mtcars
x <- "mpg"
y <- "wt"

x_plt <- rlang::sym(x)
y_plt <- rlang::sym(y)

ggplot2::ggplot(data = data, ggplot2::aes(x = !!x_plt, y = !!y_plt)) + 
    ggplot2::geom_point()

```

This approach converts the `x` and `y` string parameters to symbols and uses injection inside the `ggplot()` function.

An alternative approach is to use the `.data` pronoun (as discussed in @sec-expressions-pronouns).

```{r}
data <- mtcars
x <- "mpg"
y <- "wt"

ggplot2::ggplot(data = data, ggplot2::aes(x = .data[[x]], y = .data[[y]])) + 
    ggplot2::geom_point()

```

The pronoun approach is preferred in this instance as it is a little more concise than using symbols and works cleanly with strings.

This approach can be applied to the instruction set.  A JSON input file looks like:

```json
{
    "type": "chart",
    "mode": "scatter",
    "parameters": {
        "mapping": {
            "x": "mpg",
            "y": "wt"
        }
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", mode="scatter", parameters=list(mapping=list(x="mpg", y="wt"))), 'data/myfile.json', auto_unbox = TRUE)
```

and the engine could be:

```{r}
chart_scatter <- function(data, parameters) {
    p <- parameters$mapping
    ggplot2::ggplot(data = data, ggplot2::aes(x = .data[[p$x]], y = .data[[p$y]])) +
        ggplot2::geom_point()
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
chart_scatter(mtcars, inp$parameters)
```

ggplot2 plots support two types of aesthetics: **mapped** aesthetics, where a visual property is driven by the data, and **set** aesthetics, where a property is fixed to a constant value.  Color, for example, can be **mapped** to the value of `disp` as shown below

```json
{
    "type": "chart",
    "mode": "scatter",
    "parameters": {
        "mapping": {
            "x": "mpg",
            "y": "wt",
            "color": "disp"
        }
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", mode="scatter", parameters=list(mapping=list(x="mpg", y="wt", color="disp"))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
chart_scatter <- function(data, parameters) {
    p <- parameters$mapping
    ggplot2::ggplot(data = data, ggplot2::aes(x = .data[[p$x]], y = .data[[p$y]], color = .data[[p$color]])) +
        ggplot2::geom_point()
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
chart_scatter(mtcars, inp$parameters)
```

or set to the value `2`

```json
{
    "type": "chart",
    "mode": "scatter",
    "parameters": {
        "mapping": {
            "x": "mpg",
            "y": "wt"
        },
        "set": {
            "color": 2
        }
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", mode="scatter", parameters=list(mapping=list(x="mpg", y="wt"), set=list(color=2))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
chart_scatter <- function(data, parameters) {
    p <- parameters$mapping
    s <- parameters$set
    ggplot2::ggplot(data = data, ggplot2::aes(x = .data[[p$x]], y = .data[[p$y]])) +
        ggplot2::geom_point(color = s$color)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
chart_scatter(mtcars, inp$parameters)
```

The `ggplot()` function accepts `data` and `mapping` arguments.  The `geom_` layers can use both **mapped** and **set** aesthetics.  Mapped values in `aes()` are inherited by the layers, unless specifically overwritten.  Set values passed to the geom layer are not inherited by other layers.

The approach used for the scatter chart can be adapted to build other charts.  For example a bar chart:

```json
{
    "type": "chart",
    "mode": "bar",
    "parameters": {
        "mapping": {
            "x": "cyl"
        },
        "set": {
            "fill": 2
        }
    }
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", mode="bar", parameters=list(mapping=list(x="cyl"), set=list(fill=2))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
chart_bar <- function(data, parameters) {
    p <- parameters$mapping
    s <- parameters$set
    ggplot2::ggplot(data = data, ggplot2::aes(x = .data[[p$x]])) +
        ggplot2::geom_bar(fill = s$fill)
}

inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
chart_bar(mtcars, inp$parameters)
```

We can use a couple of the properties of ggplot to build a generic function capable of returning any chart.

  - ggplot2 geoms take the same concept of arguments (`data`, `mapping` and *other parameters*).
  - ggplot2 charts can be built up step-wise, layer by layer.

The next section will show how we can move beyond specific charts to a generic chart builder.

## A Generic Chart Builder

A generic ggplot chart builder should consider the following:

-  Properties can be inherited.  Properties defined in the global (`ggplot()`) function are inherited by graphical layers, unless specifically overwritten.  This holds true for `data` and `mapping`.
-  Properties passed to geom layers should distinguish between **mapped** and **not mapped**.
-  ggplot layers can be built up stepwise.  A chart can have multiple layers using the same or different data.

We'll start with a schema as follows:

```json
{
    "type": "chart",
    "layers": [
        {
            "type": "bar",
            "mapping": {
                "x": "cyl"
            },
            "parameters": {
                "fill": "darkgreen"
             }
        }
    ]
}
```

This schema is slightly different to the ones above in that we can specify a list of `layers` each with `mapping` and `parameter` values.

Our generic plot builder will start with a simple concept.  The `build_plot()` function can take two arguments: `data`, a rectangular data frame, possibly pre-processed through the data wrangling approaches described in @sec-data-wrangling-intro, and `params`, a list of parameters describing the geom layers.  We build a ggplot and simply add layers

```{r}
build_plot <- function(data, params) {
    
    g <- ggplot2::ggplot(data = data)
    layers <- purrr::map(params$layers, build_layer)

    for (layer in layers) {
        g <- g + rlang::eval_tidy(layer)
    }

    g
}
```

The layer is wrapped in `rlang::eval_tidy()` to evaluate it in a data mask.  This means that we can use pronouns (`.data`) to specify column names in the data.

The layers are defined using the following approach

```{r}
build_layer <- function(layer) {

    if (layer$type == "scatter") {
        geom <- ggplot2::geom_point
    } else if (layer$type == "bar") {
        geom <- ggplot2::geom_bar
    } else {
        stop("Unknown chart type ", layer$type)
    }

    ## bare geom call
    rtn <- rlang::call2(geom)

    ## add mapping
    if (!is.null(layer$mapping)) {
        mapping_exprs <- purrr::map(layer$mapping, \(x) rlang::expr(.data[[!!x]]))
        rtn[["mapping"]] <- rlang::call2(ggplot2::aes, !!!mapping_exprs)
    }

    ## add parameters that are not mapped
        for (x in names(layer$parameters)) {
            rtn[[x]] <- layer$parameters[[x]]
        }

        rtn

}
```

`build_layer()` builds and returns a single layer.  The function starts by identifying the geom and building an empty call using `rlang::call2()`.  It then adds mapping (`rlang::call2()` to `ggplot2::aes()`) and any additional parameters (passed as key-value pairs) to the geom layer before returning it.

::: {.callout-important}
### Understanding `rlang::expr(.data[[!!x]])`

`rlang::expr(.data[[!!x]])` is an important construct that allows us to build an expression of mappings.  `.data[[...]]` is a tidy-eval pronoun which tells ggplot to look up the name in the data frame at evaluation.  `!!x` injects the string (`x`) representation of a column name.  If `x` is "cyl" then `rlang::expr(.data[[!!x]])` becomes `.data[["cyl"]]`.
:::

Taking the example above:

```json
{
    "type": "chart",
    "layers": [
        {
            "type": "bar",
            "mapping": {
                "x": "cyl"
            },
            "parameters": {
                "fill": "darkgreen"
            }
        }
    ]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", layers=list(list(type="bar", mapping=list(x="cyl"), parameters=list(fill="darkgreen")))), 'data/myfile.json', auto_unbox = TRUE)
```

Running `build_plot()` returns the following:

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
build_plot(mtcars, inp)
```

### Adding `geom` Dispatcher

The concept can be extended further by including a dispatcher.  A dispatcher identifies which geom to use and returns it.

```{r}
geoms <- list(
    scatter = ggplot2::geom_point,
    bar = ggplot2::geom_bar
)
```

We can now simplify the top of `build_layer()`, removing the conditional logic used to select the geom.

```{r}
build_layer <- function(layer) {

    geom <- geoms[[layer$type]]
    if (is.null(geom)) stop("Unknown chart type ", layer$type)

    ## bare geom call
    rtn <- rlang::call2(geom)

    ## add mapping
    if (!is.null(layer$mapping)) {
        mapping_exprs <- purrr::map(layer$mapping, \(x) rlang::expr(.data[[!!x]]))
        rtn[["mapping"]] <- rlang::call2(ggplot2::aes, !!!mapping_exprs)
    }

    ## add parameters
    for (x in names(layer$parameters)) {
        rtn[[x]] <- layer$parameters[[x]]
    }

    rtn

}
```

### Default Parameters

With a slight modification to `geoms` and `build_layer()` the dispatcher can also set default options.  `build_layer()` takes the default parameters and simply adds them to the layer before any user-defined parameters are included.

```{r}
geoms <- list(
    scatter = list(
        geom = ggplot2::geom_point,
        default_params = list()
    ),
    bar = list(
        geom = ggplot2::geom_bar,
        default_params = list(width = 0.5)
    )
)
```

```{r}
build_layer <- function(layer) {

    geom <- geoms[[layer$type]]
    if (is.null(geom)) stop("Unknown chart type ", layer$type)

    ## bare geom call
    rtn <- rlang::call2(geom$geom)

    ## add mapping
    if (!is.null(layer$mapping)) {
        mapping_exprs <- purrr::map(layer$mapping, \(x) rlang::expr(.data[[!!x]]))
        rtn[["mapping"]] <- rlang::call2(ggplot2::aes, !!!mapping_exprs)
    }

    ## Default parameters
    for (x in names(geom$default_params)) {
        rtn[[x]] <- geom$default_params[[x]]
    }

    ## add parameters
    for (x in names(layer$parameters)) {
        rtn[[x]] <- layer$parameters[[x]]
    }

    rtn

}
```

```json
{
    "type": "chart",
    "layers": [
        {
            "type": "bar",
            "mapping": {
                "x": "cyl"
            },
            "parameters": {
                "fill": "darkgreen"
            }
        }
    ]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", layers=list(list(type="bar", mapping=list(x="cyl"), parameters=list(fill="darkgreen")))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
build_plot(mtcars, inp)
```

### Adding Additional Datasets

`build_plot()` is our orchestration function.  It builds the global ggplot object and adds layers to it, each layer built by `build_layer()`.  To work with multiple datasets we need to change the way `build_plot()` works, passing a named list of datasets instead of a single data frame.  Each layer can then be built using the global or a specfic dataset.

```{r}
build_plot <- function(data, params) {
    
    if (length(data) == 1) {
        global_data <- data[[1]]
    } else if ("main" %in% names(data)) {
        global_data <- data[["main"]]
    } else {
        stop("Data should a list of length 1 or a named list with a 'main' dataset")
    }

    g <- ggplot2::ggplot(data = global_data)
    layers <- purrr::map(params$layers, build_layer, data)

    for (layer in layers) {
        g <- g + rlang::eval_tidy(layer)
    }

    g
}
```

`build_layer()` can now take an additional argument `data`, which is a named list of datasets.  A layer may optionally specify a dataset to use, which overrides the global dataset for that layer only.  If no dataset is specified then the global dataset will be used.

```{r}
build_layer <- function(layer, data = NULL) {

    geom <- geoms[[layer$type]]
    if (is.null(geom)) stop("Unknown chart type ", layer$type)

    ## bare geom call
    rtn <- rlang::call2(geom$geom)

    ## add mapping
    if (!is.null(layer$mapping)) {
        mapping_exprs <- purrr::map(layer$mapping, \(x) rlang::expr(.data[[!!x]]))
        rtn[["mapping"]] <- rlang::call2(ggplot2::aes, !!!mapping_exprs)
    }

    ## data override
    ## if data included then override the global data and add it to the geom layer
    if (!is.null(layer$data) && layer$data %in% names(data)) {
        rtn[["data"]] <- data[[layer$data]]
    }

    ## Default parameters
    for (x in names(geom$default_params)) {
        rtn[[x]] <- geom$default_params[[x]]
    }

    ## add parameters
    for (x in names(layer$parameters)) {
        rtn[[x]] <- layer$parameters[[x]]
    }

    rtn

}
```

Here's an example of how this works in practice.  Let's built a plot with two layers, each scatter charts, but using different datasets.  The first layer uses the global dataset (`main`) which is simply `mtcars`.  The second layer uses a dataset called `data_set_2` which is a subset of `mtcars` where `gear == 3`.  The second layer will be plotted on top of the first layer, highlighting the points where `gear == 3` in red on top of the full dataset plotted in grey.  To achieve this we pass a named list of datasets to `build_plot()`, with `main` as the global dataset and `data_set_2` as the second dataset.  In the instruction set, layer 1 does not specify a dataset, so it uses the global dataset (`main`), while layer 2 specifies `data_set_2` as its dataset.  Because layer 2 uses a different dataset, its aesthetic mappings must be specified explicitly, even though they mirror those of layer 1.

```json
{
    "type": "chart",
    "layers": [
        {
            "type": "scatter",
            "mapping": {
                "x": "wt",
                "y": "mpg"
            },
            "parameters": {
                "color": "darkgrey"
            }
        },
        {
            "type": "scatter",
            "data": "data_set_2",
            "mapping": {
                "x": "wt",
                "y": "mpg"
            },
            "parameters": {
                "color": "red",
                "size": 4,
                "shape": 1
            }
        }
    ]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", layers=list(list(type="scatter", mapping=list(x="wt", y="mpg"), parameters=list(color="darkgrey")), list(type="scatter", data="data_set_2", mapping=list(x="wt", y="mpg"), parameters=list(color="red", size=4, shape=1)))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
df <- dplyr::filter(mtcars, gear == 3)
data_list <- list(main = mtcars, data_set_2 = df)
build_plot(data_list, inp)
```

### Extensibility

This approach is highly extensible.  To add a new chart, simply add a new term to `geoms`.  For example

```{r}
geoms <- list(
    scatter = list(
        geom = ggplot2::geom_point,
        default_params = list()
    ),
    bar = list(
        geom = ggplot2::geom_bar,
        default_params = list(width = 0.5)
    ),
    boxplot = list(
        geom = ggplot2::geom_boxplot
    )
)
```

```json
{
    "type": "chart",
    "layers": [
        {
            "type": "boxplot",
            "mapping": {
                "x": "cyl",
                "group": "cyl",
                "y": "mpg"
            },
            "parameters": {
                "fill": "darkgreen"
            }
        }
    ]
}
```

```{r}
#| echo: false
#| eval: true
jsonlite::write_json(list(type="chart", layers=list(list(type="boxplot", mapping=list(x="cyl", group="cyl", y="mpg"), parameters=list(fill="darkgreen")))), 'data/myfile.json', auto_unbox = TRUE)
```

```{r}
inp <- jsonlite::fromJSON("data/myfile.json", simplifyVector = FALSE)
build_plot(list(mtcars), inp)
```
